---
title: Evaluating diagnostic </br> classification models
subtitle: With Stan and measr
author: W. Jake Thompson, Ph.D.
---

```{r setup, include = FALSE}
library(tidyverse)
library(ggmeasr)
library(ggdist)
library(magick)
library(distributional)
library(knitr)
library(measr)
library(here)
library(fs)
library(downlit)
library(countdown)
library(cmdstanr)
library(posterior)

opts_chunk$set(
  fig.width = 7,
  fig.asp = 0.618,
  fig.align = "center"
)

set_theme(plot_margin = margin(5, 0, 0, 0))

# copy grid-worms
if (!dir_exists(here("_site/materials/slides/"))) {
  dir_create(here("_site/materials/slides/"))
}
dir_copy(
  here("materials", "slides", "grid-worms"),
  here("_site/materials/slides/grid-worms"),
  overwrite = TRUE
)
```

## Model evaluation {background-image="figure/backgrounds/default.png" background-size="contain"}

* **Absolute fit**: How well does the model represent the observed data?
  * Model-level
  * Item-level

* **Relative fit**: How do multiple models compare to each other?

* **Reliability**: How consistent and accurate are the classifications?

# Absolute model fit {background-image="figure/backgrounds/section.png" background-size="contain"}

## Model-level absolute fit {background-image="figure/backgrounds/default.png" background-size="contain"}

* Limited information indices based on parameter point estimates
  * M~2~ statistic ([Liu et al., 2016](https://doi.org/10.3102/1076998615621293))
  
* Posterior predictive model checks (PPMCs)

## Calculating limited information indices {background-image="figure/backgrounds/default.png" background-size="contain"}

```{r read-ecpe}
ecpe <- measr_dcm(
  data = ecpe_data,
  qmatrix = ecpe_qmatrix,
  resp_id = "resp_id",
  item_id = "item_id",
  type = "lcdm",
  method = "mcmc",
  backend = "cmdstanr",
  iter_warmup = 1000,
  iter_sampling = 500,
  chains = 4,
  parallel_chains = 4,
  file = "fits/ecpe-lcdm"
)
```

:::{.panel-tabset}

### One-time calculation

`fit_*` functions are used for calculating absolute fit indices.

```{r ecpe-m2}
#| echo: true
fit_m2(ecpe)
```

### Save for future

`add_fit()` will add the fit index to the model object (and resave) to prevent unnecessary calculations in the future.

```{r ecpe-add-m2}
#| echo: true

ecpe <- add_fit(ecpe, method = "m2")
measr_extract(ecpe, "m2")
```

:::

## Exercise 1 {.exercise background-image="figure/backgrounds/exercises.png" background-size="contain"}

* Open `evaluation.Rmd` and run the `setup` chunk.

* Calculate the M~2~ statistic for the Taylor LCDM model using `add_fit()`.

* Extract the M~2~ statistic. Does the model fit the data?

```{r countdown-1}
countdown(
  minutes = 3,
  seconds = 0,
  bottom = "10%",
  color_border = "#023047",
  color_background = "#023047",
  color_text = "#8ECAE6",
  color_running_background = "#023047",
  color_running_text = "#8ECAE6",
  color_finished_background = "#D7263D",
  color_finished_text = "#FFFFFF"
)
```

## {.exercise data-menu-title="Solution 1" background-image="figure/backgrounds/exercises.png" background-size="contain"}

```{r read-taylor}
taylor_data <- read_rds(here("materials", "slides", "data", "taylor-data.rds"))
taylor_qmatrix <- read_rds(here(
  "materials",
  "slides",
  "data",
  "taylor-qmatrix.rds"
))

taylor_lcdm <- measr_dcm(
  data = taylor_data,
  qmatrix = taylor_qmatrix,
  resp_id = "album",
  type = "lcdm",
  method = "mcmc",
  backend = "rstan",
  warmup = 1000,
  iter = 1500,
  chains = 2,
  cores = 2,
  file = "fits/taylor-lcdm"
)
```

```{r taylor-m2, echo = TRUE}
taylor_lcdm <- add_fit(taylor_lcdm, method = "m2")
measr_extract(taylor_lcdm, "m2")
```

* The null hypothesis is that our model fits the data
  * *p*-values > .05 indicate acceptable model fit

## PPMC: Raw score distribution {background-image="figure/backgrounds/default.png" background-size="contain"}

```{r calc-raw-score, eval = !file.exists(here("materials", "slides", "data", "ecpe-raw-scores.rds"))}
retain <- 500
keep_draws <- sample(seq_len(ndraws(as_draws(ecpe))), size = retain)
ecpe_results <- predict(ecpe, summary = FALSE)

pi <- as_draws_df(ecpe) |>
  merge_chains() |>
  subset_draws(variable = "pi", draw = keep_draws) |>
  as_tibble() |>
  pivot_longer(starts_with("pi")) |>
  separate_wider_regex(
    name,
    patterns = c("pi\\[", item = "[0-9]*", ",", class = "[0-9]*", "\\]")
  ) |>
  select(.draw, item, class, prob = value)

format_draws <- function(x, keep) {
  ret <- as_draws_df(x) |>
    merge_chains() |>
    subset_draws(draw = keep) |>
    as_tibble() |>
    select(.draw, starts_with("x")) |>
    pivot_longer(cols = -.draw, names_to = "resp_id", values_to = "prob")

  return(ret)
}

all_draws <- vector(
  mode = "list",
  length = ncol(ecpe_results$class_probabilities) - 1
)
for (i in seq_len(ncol(ecpe_results$class_probabilities))[-1]) {
  cur_name <- colnames(ecpe_results$class_probabilities)[i]
  ret <- format_draws(
    ecpe_results$class_probabilities[[i]],
    keep = keep_draws
  ) |>
    mutate(
      resp_id = str_replace(resp_id, "x\\[([0-9]*)\\]", "\\1"),
      resp_id = as.integer(resp_id)
    ) |>
    rename(!!cur_name := prob)

  all_draws[[i - 1]] <- ret
}

raw_scores <- all_draws |>
  reduce(full_join, join_by(.draw, resp_id)) |>
  pivot_longer(starts_with("[")) |>
  group_by(.draw, resp_id) |>
  mutate(class = 1:n()) |>
  slice_sample(n = 1, weight_by = value) |>
  ungroup() |>
  select(.draw, resp_id, class) |>
  mutate(.draw = as.integer(factor(.draw)), class = as.character(class)) |>
  left_join(pi, join_by(.draw, class), relationship = "many-to-many") |>
  mutate(rand = runif(n()), score = as.integer(rand <= prob)) |>
  summarize(score = sum(score), .by = c(.draw, resp_id)) |>
  count(.draw, score)

write_rds(
  raw_scores,
  here("materials", "slides", "data", "ecpe-raw-scores.rds")
)
```

```{r}
raw_scores <- read_rds(here(
  "materials",
  "slides",
  "data",
  "ecpe-raw-scores.rds"
))
```

:::{.columns}

:::{.column width="30%"}
* For each iteration, calculate the total number of respondents at each score point

:::{.fragment fragment-index=2}

* Calculate the expected number of respondents at each score point

:::

:::{.fragment fragment-index=3}

* Calculate the <span style="color: #D7263D;">observed</span> number of respondents at each score point

:::

:::

:::{.column width="70%"}

:::{.r-stack}

:::{.fragment fragment-index=1}

```{r score-dist}
#| out-width: 100%
#| out-height: 50%
#| fig-alt: "Scatter plot showing the number of respondents at each score point in each iteration."

p <- ggplot() +
  geom_point(
    data = raw_scores,
    aes(x = factor(score), y = n),
    position = position_jitter(height = 0, seed = 1213),
    alpha = 0.2,
    color = palette_measr[4]
  ) +
  scale_y_comma() +
  labs(x = "Correct Responses", y = "Respondents")

p
```

:::

:::{.fragment fragment-index=2}

```{r exp-score}
#| out-width: 100%
#| out-height: 50%
#| fig-alt: "Scatter plot showing the number of respondents at each score point in each iteration with the average number of respondents overlayed."

exp_scores <- summarize(raw_scores, n = mean(n), .by = score)

p <- p +
  geom_point(
    data = exp_scores,
    aes(x = factor(score), y = n),
    color = palette_measr[1],
    shape = 18,
    size = 5
  ) +
  geom_line(
    data = exp_scores,
    aes(x = factor(score), y = n),
    group = 1,
    color = palette_measr[1]
  )

p
```

:::

:::{.fragment fragment-index=3}

```{r obs-score}
#| out-width: 100%
#| out-height: 50%
#| fig-alt: "Scatter plot showing the number of respondents at each score point in each iteration with the average and observed number of respondents overlayed."

obs_scores <- ecpe_data |>
  pivot_longer(-resp_id) |>
  summarize(score = sum(value), .by = resp_id) |>
  count(score)

p <- p +
  geom_point(
    data = obs_scores,
    aes(x = factor(score), y = n),
    color = palette_measr[2],
    shape = 16,
    size = 5
  ) +
  geom_line(
    data = obs_scores,
    aes(x = factor(score), y = n),
    color = palette_measr[2],
    group = 1
  )

p
```

:::

:::

:::

:::

## PPMC: $\chi^2$ {background-image="figure/backgrounds/default.png" background-size="contain"}

:::{.columns}

:::{.column width="30%"}
* Calculate a $\chi^2$-like statistic comparing the number of respondents at each score point in each iteration to the expectation

:::{.fragment fragment-index=2}

* Calculate the $\chi^2$ value comparing the <span style="color: #D7263D;">observed</span> data to the expectation

:::

:::

:::{.column width="70%"}

:::{.r-stack}

:::{.fragment fragment-index=1}

```{r chisq-dist}
#| out-width: 100%
#| out-height: 50%
#| fig-alt: "Histogram of the chi-square values from each iteration."

ppmc_chisq <- raw_scores |>
  complete(.draw, score, fill = list(n = 0L)) |>
  full_join(rename(exp_scores, exp = n), join_by(score)) |>
  arrange(.draw, score) |>
  mutate(piece = ((n - exp)^2) / exp) |>
  summarize(chisq = sum(piece), .by = .draw)

p <- ggplot() +
  geom_histogram(
    data = ppmc_chisq,
    aes(x = chisq),
    binwidth = 2,
    boundary = 0,
    fill = palette_measr[1],
    color = palette_measr[4]
  ) +
  labs(x = "&chi;<sup>2</sup><sub>rep</sub>", y = "Replications") +
  theme(axis.title.x = ggtext::element_markdown(family = "sans"))

p
```

:::

:::{.fragment fragment-index=2}

```{r chisq-obs}
#| out-width: 100%
#| out-height: 50%
#| fig-alt: "Histogram of the chi-square values from each iteration with a dashed vertical line indicating the value from the observed data."

obs_chisq <- obs_scores |>
  full_join(rename(exp_scores, exp = n), join_by(score)) |>
  replace_na(list(n = 0L)) |>
  arrange(score) |>
  mutate(piece = ((n - exp)^2) / exp) |>
  summarize(chisq = sum(piece))

p <- p +
  geom_vline(
    xintercept = obs_chisq$chisq,
    linetype = "dashed",
    color = palette_measr[2]
  )

p
```

:::

:::

:::

:::

## PPMC: *ppp* {background-image="figure/backgrounds/default.png" background-size="contain"}

* Calculate the proportion of iterations where the $\chi^2$-like statistic from replicated data set exceeds the observed data statistic
  * Posterior predictive *p*-value (*ppp*)
  
* Very high values (e.g., >.975) or very low values (e.g., <.025) indicate our observed value is far in the tails of what the model would predict
  * Poor model fit

* Ideally, we'd like *ppp* values close to .5

## PPMCs with measr {background-image="figure/backgrounds/default.png" background-size="contain"}

:::{.panel-tabset}

### One-time calculation

```{r ecpe-raw-score}
#| echo: true
fit_ppmc(ecpe, model_fit = "raw_score", item_fit = NULL)
```

### Save for future

```{r ecpe-save-raw-score}
#| echo: true

ecpe <- add_fit(ecpe, method = "ppmc", model_fit = "raw_score", item_fit = NULL)
measr_extract(ecpe, "ppmc_raw_score")
```

:::

## Exercise 2 {.exercise background-image="figure/backgrounds/exercises.png" background-size="contain"}

* Calculate the raw score PPMC for the Taylor LCDM

* Does the model fit the observed data?

```{r countdown-2}
countdown(minutes = 4, bottom = "10%")
```

## {.exercise data-menu-title="Solution 2" background-image="figure/backgrounds/exercises.png" background-size="contain"}

```{r taylor-raw-score, echo = TRUE}
taylor_lcdm <- add_fit(taylor_lcdm, method = "ppmc", item_fit = NULL)
measr_extract(taylor_lcdm, "ppmc_raw_score")
```

* The *ppp* value is between .025 and .975
* Model fit appears to be adequate.

## Item-level fit {background-image="figure/backgrounds/default.png" background-size="contain"}

* Diagnose problems with model-level

* Identify particular items that may not be performing as expected

* Identify potential dimensionality issues

## Item-level fit with measr {background-image="figure/backgrounds/default.png" background-size="contain"}

* Currently support two-measures of item-level fit using PPMCs:
  * Conditional probability of each class providing a correct response ($\pi$ matrix)
  * Item pair odds ratios

## Calculating item-level fit {background-image="figure/backgrounds/default.png" background-size="contain"}

:::{.panel-tabset}

### One-time calculation

```{r ecpe-odds-ratio}
#| echo: true
fit_ppmc(ecpe, model_fit = NULL, item_fit = "odds_ratio")
```

### Save for future

```{r ecpe-save-ppmc-item}
#| echo: true

ecpe <- add_fit(
  ecpe,
  method = "ppmc",
  model_fit = NULL,
  item_fit = c("conditional_prob", "odds_ratio")
)
```

:::


## Extracting item-level fit {background-image="figure/backgrounds/default.png" background-size="contain"}

:::{.panel-tabset}

### Conditional probabilities

```{r ecpe-cond-prob, echo = TRUE}
measr_extract(ecpe, "ppmc_conditional_prob")
```

### Odds ratios

```{r ecpe-odd-ratio, echo = TRUE}
measr_extract(ecpe, "ppmc_odds_ratio")
```

:::

## Flagging item-level fit {background-image="figure/backgrounds/default.png" background-size="contain"}

:::{.panel-tabset}

### Conditional probabilities

```{r ecpe-cond-prob-flags, echo = TRUE}
measr_extract(ecpe, "ppmc_conditional_prob_flags")
```

### Odds ratios

```{r ecpe-odd-ratio-flags, echo = TRUE}
measr_extract(ecpe, "ppmc_odds_ratio_flags")
```

:::

## Patterns of item-level misfit {background-image="figure/backgrounds/default.png" background-size="contain"}

```{r flag-patterns, echo = TRUE}
#| class-output: highlight
#| output-line-numbers: "|4-7"

measr_extract(ecpe, "ppmc_conditional_prob_flags") |>
  count(class, name = "flags") |>
  left_join(measr_extract(ecpe, "strc_param"), by = join_by(class)) |>
  arrange(desc(flags))
```

## Exercise 3 {.exercise background-image="figure/backgrounds/exercises.png" background-size="contain"}

* Calculate PPMCs for the conditional probabilities and odds ratios for the Taylor model

* What do the results tell us about the model?

```{r countdown-3}
countdown(minutes = 5, bottom = "10%")
```

## {.exercise data-menu-title="Solution 3" background-image="figure/backgrounds/exercises.png" background-size="contain"}

```{r taylor-item-fit, echo = TRUE}
taylor_lcdm <- add_fit(
  taylor_lcdm,
  method = "ppmc",
  item_fit = c("conditional_prob", "odds_ratio")
)
```

</br>

:::{.panel-tabset}

### Conditional probabilities

```{r taylor-cond-prob, echo = TRUE}
measr_extract(taylor_lcdm, "ppmc_conditional_prob_flags")
```

### Odds ratios

```{r taylor-odd-ratio, echo = TRUE}
measr_extract(taylor_lcdm, "ppmc_odds_ratio_flags")
```

:::

# Relative model fit {background-image="figure/backgrounds/section.png" background-size="contain"}

## Model comparisons  {background-image="figure/backgrounds/default.png" background-size="contain"}

* Doesn't give us information whether or not a model fits the data, only compares competing models to each other
  * Should be evaluated in conjunction with absolute model fit

* Implemented with the [loo](https://mc-stan.org/loo) package
  * PSIS-LOO ([Vehtari, 2017](https://doi.org/10.1007/s11222-016-9696-4))
  * WAIC ([Watanabe, 2010](https://jmlr.org/papers/v11/watanabe10a.html))

## Relative fit with measr {background-image="figure/backgrounds/default.png" background-size="contain"}

```{r ecpe-rel-fit, echo = TRUE}
ecpe <- add_criterion(ecpe, criterion = c("loo", "waic"))
```

</br>

:::{.fragment}

```{r ecpe-loo, echo = TRUE}
measr_extract(ecpe, "loo")
```

:::

## Comparing models {background-image="figure/backgrounds/default.png" background-size="contain"}

First, we need another model to compare

```{r ecpe-dina, echo = TRUE}
#| code-line-numbers: "|4|11"

ecpe_dina <- measr_dcm(
  data = ecpe_data,
  qmatrix = ecpe_qmatrix,
  resp_id = "resp_id",
  item_id = "item_id",
  type = "dina",
  method = "mcmc",
  backend = "cmdstanr",
  iter_warmup = 1000,
  iter_sampling = 500,
  chains = 4,
  parallel_chains = 4,
  file = "fits/ecpe-dina"
)

ecpe_dina <- add_criterion(ecpe_dina, criterion = "loo")
```

## `loo_compare()` {background-image="figure/backgrounds/default.png" background-size="contain"}

```{r compare-loo, echo = TRUE}
loo_compare(ecpe, ecpe_dina, criterion = "loo", model_names = c("lcdm", "dina"))
```

* LCDM is the preferred model
  * Preferred does not imply "good"
  * Remember, the LCDM showed poor absolute fit

* Difference is much larger than the standard error of the difference
  * Approximate threshold: Absolute difference is greater than 2.5 &times; standard error of the difference ([Bengio & Grandvalet, 2004](https://www.jmlr.org/papers/v5/grandvalet04a.html))

## Exercise 4 {.exercise background-image="figure/backgrounds/exercises.png" background-size="contain"}

* Estimate a DINA model for the Taylor data

* Add PSIS-LOO and WAIC criteria to both the LCDM and DINA models for the Taylor data

* Use `loo_compare()` to compare the LCDM and DINA models
  * What do the findings tell us?
  * Can you explain the results?

```{r countdown-4}
countdown(minutes = 15, bottom = "10%")
```

## {.exercise data-menu-title="Solution 4" background-image="figure/backgrounds/exercises.png" background-size="contain"}

:::{.panel-tabset}

### Estimate DINA model

```{r taylor-dina, echo = TRUE}
taylor_dina <- measr_dcm(
  data = taylor_data,
  qmatrix = taylor_qmatrix,
  resp_id = "album",
  type = "dina",
  method = "mcmc",
  backend = "rstan",
  warmup = 1000,
  iter = 1500,
  chains = 2,
  cores = 2,
  file = "fits/taylor-dina"
)
```

### Add criterion

```{r taylor-rel-fit, echo = TRUE}
taylor_lcdm <- add_criterion(taylor_lcdm, criterion = c("loo", "waic"))
taylor_dina <- add_criterion(taylor_dina, criterion = c("loo", "waic"))
```

### Compare models


```{r taylor-compare-loo, echo = TRUE}
loo_compare(taylor_lcdm, taylor_dina, criterion = "loo")
```

</br>

```{r taylor-compare-waic, echo = TRUE}
loo_compare(taylor_lcdm, taylor_dina, criterion = "waic")
```

:::

# Reliability {background-image="figure/backgrounds/section.png" background-size="contain"}

## Reliability methods {background-image="figure/backgrounds/default.png" background-size="contain"}

* Reporting reliability depends on how results are estimated and reported

* Reliability for:
  * Profile-level classification
  * Attribute-level classification
  * Attribute-level probability of proficiency

## Reliability with measr {background-image="figure/backgrounds/default.png" background-size="contain"}

:::{.panel-tabset}

### One-time calculation

```{r ecpe-reli, cache = TRUE}
#| echo: true
reliability(ecpe)
```

### Save for future

```{r ecpe-add-reli}
#| echo: true

ecpe <- add_reliability(ecpe)
```

:::

## Profile-level classification {background-image="figure/backgrounds/default.png" background-size="contain"}

:::{.panel-tabset}

### Profile-level probabilities

```{r ecpe-class-prob, echo = TRUE}
measr_extract(ecpe, "class_prob")
```

### Profile-level classifications

:::{.columns}

:::{.column width="60%"}

```{r ecpe-class-show, echo = TRUE, eval = FALSE}
measr_extract(ecpe, "class_prob") |>
  pivot_longer(-resp_id, names_to = "profile", values_to = "prob") |>
  slice_max(order_by = prob, by = resp_id)
```

:::

:::{.column width="40%"}

```{r ecpe-class-eval, echo = FALSE, eval = TRUE}
measr_extract(ecpe, "class_prob") |>
  pivot_longer(-resp_id, names_to = "profile", values_to = "prob") |>
  slice_max(order_by = prob, by = resp_id)
```

:::

:::

### Profile reliability

```{r ecpe-class-reli, echo = TRUE}
measr_extract(ecpe, "pattern_reliability")
```

* Estimating classification consistency and accuracy for cognitive diagnostic assessment ([Cui et al., 2012](https://doi.org/10.1111/j.1745-3984.2011.00158.x))

:::

## Attribute-level classification {background-image="figure/backgrounds/default.png" background-size="contain"}

:::{.panel-tabset}

### Attribute-level probabilities

```{r ecpe-att-prob, echo = TRUE}
measr_extract(ecpe, "attribute_prob")
```

### Attribute-level classifications

:::{.columns}

:::{.column width="45%"}

```{r ecpe-att-show, echo = TRUE, eval = FALSE}
measr_extract(ecpe, "attribute_prob") |>
  mutate(
    across(
      where(is.double),
      ~ case_when(.x > 0.5 ~ 1L, TRUE ~ 0L)
    )
  )
```

:::

:::{.column width="55%"}

```{r ecpe-att-eval, echo = FALSE, eval = TRUE}
measr_extract(ecpe, "attribute_prob") |>
  mutate(
    across(
      where(is.double),
      ~ case_when(.x > 0.5 ~ 1L, TRUE ~ 0L)
    )
  )
```

:::

:::

### Classification reliability

```{r ecpe-att-reli, echo = TRUE}
measr_extract(ecpe, "classification_reliability")
```

* Measures of agreement to assess attribute-level classification accuracy and consistency for cognitive diagnostic assessments ([Johnson & Sinharay, 2018](https://doi.org/10.1111/jedm.12196))

:::

## Attribute-level probabilities {background-image="figure/backgrounds/default.png" background-size="contain"}

:::{.panel-tabset}

### Attribute-level probabilities

```{r ecpe-att-prob-2, echo = TRUE}
measr_extract(ecpe, "attribute_prob")
```

### Probability reliability

```{r ecpe-prob-reli, echo = TRUE}
measr_extract(ecpe, "probability_reliability")
```

* The reliability of the posterior probability of skill attainment in diagnostic classification models ([Johnson & Sinharay, 2020](https://doi.org/10.3102/1076998619864550))

:::

## Exercise 5 {.exercise background-image="figure/backgrounds/exercises.png" background-size="contain"}

* Add reliability information to the Taylor LCDM and DINA models

* Examine the attribute classification indices for both models

```{r countdown-5}
countdown(minutes = 4, bottom = "10%")
```

## {.exercise data-menu-title="Solution 5" background-image="figure/backgrounds/exercises.png" background-size="contain"}

:::{.panel-tabset}

### Calculate reliability

```{r taylor-calc-reli, echo = TRUE}
taylor_lcdm <- add_reliability(taylor_lcdm)
taylor_dina <- add_reliability(taylor_dina)
```

### Compare reliability

```{r taylor-lcdm-reli, echo = TRUE}
measr_extract(taylor_lcdm, "classification_reliability")
```

</br>

```{r taylor-dina-reli, echo = TRUE}
measr_extract(taylor_dina, "classification_reliability")
```

:::

# {.closing data-menu-title="Closing" background-image="figure/backgrounds/plain-title.png" background-size="contain" style="top: 0px;"}

</br>

:::{.end-title color="white" font-size="200%"}
Evaluating diagnostic classification models
:::

:::{.end-subtitle}
With Stan and measr
:::

:::{.center}
<https://ncme2024.measr.info>
:::
