---
title: Diagnostic classification <br> models
subtitle: A brief introduction
author: W. Jake Thompson, Ph.D.
---

# What are diagnostic models?

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(ggmeasr)
library(ggdist)
library(magick)
library(distributional)
library(knitr)
library(measr)
library(here)

set_theme(plot_margin = margin(5, 0, 0, 0))
```

## {data-menu-title="Our Example"}

```{r}
#| label: taylor-grammys
#| out-width: 100%
#| fit-alt: "Artistic renderings of Taylor Swift after her Grammy wins."

include_graphics(here(
  "materials",
  "slides",
  "figure",
  "taylor",
  "taylor-grammys.png"
))
```

## {data-menu-title="Taylor's Eras" .empty-white}

```{r}
#| label: all-taylor
#| out-width: 100%
#| fit-alt: "Artistic renderings of Taylor Swift from all 13 album releases."

include_graphics(here(
  "materials",
  "slides",
  "figure",
  "taylor",
  "all-taylor.png"
))
```

## {data-menu-title="Classic psychometrics"}

:::{.columns}
:::{.column width="25%"}
* Traditional assessments and psychometric models measure an overall skill or ability
* Assume a continuous latent trait
:::

:::{.column width="75%"}
```{r}
#| label: taylor-continuum
#| out-width: 100%
#| out-height: 90%
#| fig-alt: |
#|   A normal distribution with images of Taylor Swift from each era overlayed.

taylors <- read_rds(here(
  "materials",
  "slides",
  "data",
  "taylor-results.rds"
)) |>
  mutate(
    x = theta,
    y = 0.02,
    img = here("materials", "slides", "figure", "taylor", "eras", img)
  )

base <- prior(normal(0, 1), type = "intercept") |>
  dcmstan::prior_tibble() |>
  parse_dist(prior) |>
  ggplot(aes(xdist = .dist_obj)) +
  stat_slab(color = palette_measr[1], fill = palette_measr[3])

width <- 0.6
for (i in 1:nrow(taylors)) {
  img <- as.raster(image_read(taylors$img[i]))

  base <- base +
    annotation_raster(
      img,
      taylors$x[i] - (width / 2),
      taylors$x[i] + (width / 2),
      taylors$y[i],
      taylors$y[i] + 0.25
    )
}

base +
  labs(x = "Musical Knowledge", y = NULL) +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank())
```

:::
:::

## {data-menu-title="Traditional measurement: Weak ordering"}

:::{.columns}
:::{.column width="25%"}
* The output is a weak ordering due to error in estimates
  * Confident *Taylor Swift* (debut) is the worst
  * Not confident on ordering toward the middle of the distribution
:::

:::{.column width="75%"}
```{r}
#| label: taylor-continuum
#| out-width: 100%
#| out-height: 90%
#| fig-alt: |
#|   A normal distribution with images of Taylor Swift from each era overlayed.
```

:::
:::

## {data-menu-title="Traditional measurement: Student feedback"}

:::{.columns}
:::{.column width="25%"}
* Limited in the types of questions that can be answered. 
  * Why is *Taylor Swift* (debut) so low?
  * What aspects do each era demonstrate proficiency or competency of?
  * How much skill is "enough" to be competent?
:::

:::{.column width="75%"}
```{r}
#| label: taylor-continuum
#| out-width: 100%
#| out-height: 90%
#| fig-alt: |
#|   A normal distribution with images of Taylor Swift from each era overlayed.
```

:::
:::
  
## Diagnostic measurement

* Designed to be multidimensional
* No continuum of student achievement
* Categorical constructs
  * Usually binary (e.g., master/nonmaster, proficient/not proficient)

* Several different names in the literature
  * Diagnostic classification models (DCMs)
  * Cognitive diagnostic models (CDMs)
  * Skills assessment models
  * Latent response models
  * Restricted latent class models

## Diagnostic music assessment

:::{.columns}

:::{.column width="30%"}
* Rather than measuring overall musical knowledge, we can break music down into set of skills or *attributes*
  * Songwriting
  * Production
  * Vocals
:::

:::{.column width="70%"}
```{r}
#| label: music-attributes
#| fig-asp: 0.3
#| out-width: 80%
#| out-height: 40%
#| fig-alt: |
#|   Three circles representing the 3 attributes. The bottom half of each circle
#|   is shaded dark, and the top half is light, to indicate there are two
#|   categories for each attribute.

library(ggforce)

tibble(
  start = rep(c(-pi / 2, pi / 2), 3),
  type = rep(c("Proficient", "Non-proficient"), 3),
  skill = rep(c("Songwriting", "Production", "Vocals"), each = 2),
  x = rep(c(3, 6, 9), each = 2),
  y = 0
) |>
  ggplot() +
  geom_arc_bar(
    aes(
      x0 = x,
      y0 = y,
      r0 = 0,
      r = 1.2,
      start = start,
      end = start + pi,
      fill = type
    ),
    color = "white",
    show.legend = FALSE,
    radius = 0
  ) +
  geom_text(
    data = ~ slice(., c(1, 3, 5, 7)),
    aes(x = x, y = 0.2, label = skill),
    size = 5
  ) +
  scale_fill_manual(
    values = c(
      "Proficient" = palette_measr[3],
      "Non-proficient" = palette_measr[1]
    )
  ) +
  coord_equal() +
  theme_void()
```

:::
:::

* Attributes are categorical, often dichotomous (e.g., proficient vs. non-proficient)

## Diagnostic classification models

:::{.columns}
:::{.column width="50%"}
* DCMs place individuals into groups according to proficiency of multiple attributes
:::

:::{.column width="50%"}
```{r}
#| label: taylor-profiles

library(gt)
library(gtExtras)

taylor_profiles <- taylors |>
  mutate(across(c(songwriting, production, vocals), \(x) {
    case_when(x == 1 ~ "check", x == 0 ~ "xmark")
  })) |>
  mutate(
    s_color = case_when(
      songwriting == "xmark" ~ palette_measr[2],
      songwriting == "check" ~ palette_measr[4]
    ),
    p_color = case_when(
      production == "xmark" ~ palette_measr[2],
      production == "check" ~ palette_measr[4]
    ),
    v_color = case_when(
      vocals == "xmark" ~ palette_measr[2],
      vocals == "check" ~ palette_measr[4]
    )
  ) |>
  select(
    era,
    album_release,
    img,
    songwriting,
    production,
    vocals,
    ends_with("_color")
  )

taylor_profiles |>
  filter(era %in% c("Debut", "1989", "reputation", "folklore")) |>
  select(img, songwriting, production, vocals, ends_with("_color")) |>
  gt() |>
  cols_hide(ends_with("_color")) |>
  cols_label(img = "") |>
  # cols_width(img ~ px(100),
  #            everything() ~ px(150)) |>
  gt_img_rows(columns = img, img_source = "local", height = 75) |>
  fmt_icon(songwriting, fill_color = from_column("s_color"), height = "40px") |>
  fmt_icon(production, fill_color = from_column("p_color"), height = "40px") |>
  fmt_icon(vocals, fill_color = from_column("v_color"), height = "40px") |>
  cols_align("center", -img) |>
  gt_theme_measr() |>
  tab_options(table.font.size = 24)
```

:::
:::

## Benefits of DCMs

:::{.columns}
:::{.column width="50%"}
* Fine-grained, multidimensional results. Answer more questions:
  * Why is *Taylor Swift* (debut) so low?
    * Subpar songwriting, production, and vocals
  * What aspects are albums competent/proficient in?
    * DCMs provide classifications directly

* High reliability with fewer items
  * Less information need to classify than to place precisely along a scale
:::

:::{.column width="50%"}
```{r}
#| label: taylor-profiles
```

:::
:::


## Results from DCM-based assessments

:::{.columns}

:::{.column width="50%"}

```{r}
#| label: all-profiles

taylor_profiles |>
  arrange(album_release) |>
  select(img, songwriting, production, vocals, ends_with("_color")) |>
  gt() |>
  cols_hide(ends_with("_color")) |>
  cols_label(img = "") |>
  # cols_width(img ~ px(100),
  #            everything() ~ px(150)) |>
  gt_img_rows(columns = img, img_source = "local", height = 75) |>
  fmt_icon(songwriting, fill_color = from_column("s_color"), height = "40px") |>
  fmt_icon(production, fill_color = from_column("p_color"), height = "40px") |>
  fmt_icon(vocals, fill_color = from_column("v_color"), height = "40px") |>
  cols_align("center", -img) |>
  gt_theme_measr() |>
  tab_options(
    table.font.size = 18,
    container.height = px(500),
    container.overflow.y = TRUE
  )
```

:::

:::{.column width="50%"}

* No scale, no overall "ability"
* Students are probabilistically placed into classes
  * Classes are represented by skill profiles
* Feedback on specific skills as defined by the cognitive theory and test design

:::

:::

## Fine-grained feedback

* Distinguish between respondents who may have similar scale scores

:::{.columns}
:::{.column width="50%"}
```{r}
#| label: mid-profiles

taylor_profiles |>
  semi_join(
    taylors |>
      filter(between(theta, -0.5, 0.5)),
    join_by(era)
  ) |>
  select(img, songwriting, production, vocals, ends_with("_color")) |>
  gt() |>
  cols_hide(ends_with("_color")) |>
  cols_label(img = "") |>
  # cols_width(img ~ px(100),
  #            everything() ~ px(150)) |>
  gt_img_rows(columns = img, img_source = "local", height = 75) |>
  fmt_icon(songwriting, fill_color = from_column("s_color"), height = "40px") |>
  fmt_icon(production, fill_color = from_column("p_color"), height = "40px") |>
  fmt_icon(vocals, fill_color = from_column("v_color"), height = "40px") |>
  cols_align("center", -img) |>
  gt_theme_measr() |>
  tab_options(
    table.font.size = 18,
    container.height = px(400),
    container.overflow.y = TRUE
  )
```

:::

:::{.column width="50%"}
```{r}
#| label: mid-scale

table_taylors <- taylors |>
  filter(between(theta, -0.5, 0.5))

prior(normal(0, 1), type = "intercept") |>
  dcmstan::prior_tibble() |>
  parse_dist(prior) |>
  ggplot(aes(xdist = .dist_obj)) +
  stat_slab(color = palette_measr[1], fill = palette_measr[3]) -> base

width <- 0.6
for (i in 1:nrow(table_taylors)) {
  img <- as.raster(image_read(table_taylors$img[i]))

  base <- base +
    annotation_raster(
      img,
      table_taylors$x[i] - (width / 2),
      table_taylors$x[i] + (width / 2),
      table_taylors$y[i],
      table_taylors$y[i] + 0.25
    )
}

base +
  labs(x = "Musical Knowledge", y = NULL) +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank())
```

:::
:::

## Item structures for DCMs

:::{.columns}

:::{.column width="60%"}

* Item structure: Which skills are measured by each item?
  * Simple structure: Item measures a single skill
  * Complex structure: Item measures 2+ skills

* Defined by Q-matrix

* Interactions between attributes when an item measures multiple skills driven by cognitive theory and/or empirical evidence
  * Can proficiency of one skill compensate for non-proficiency of another?
  * Are skill acquired in a particular order (e.g., hierarchy)?

:::

:::{.column width="40%"}

```{r}
#| label: taylor-qmatrix

read_rds(here("materials", "slides", "data", "taylor-qmatrix.rds")) |>
  rowid_to_column(var = "item") |>
  gt() |>
  gt_theme_measr() |>
  cols_align(align = "center") |>
  tab_options(
    table.font.size = 18,
    container.height = px(500),
    container.overflow.y = TRUE
  )
```

:::

:::

## Classification reliability

:::{.columns}

:::{.column width="30%"}

* Easier to categorize than place along a continuum

:::{.fragment fragment-index=3}
* Can set a proficiency threshold to optimize Type 1 or Type 2 errors
:::

:::

:::{.column width="70%"}
:::{.r-stack}
:::{.fragment fragment-index=1}

```{r}
#| label: theta-reliability
#| warning: false
#| out-width: 100%
#| out-height: 100%
#| fig-alt: "Line graph showing a normal distribution with a peak around 1.5."

tibble(x = seq(-3, 3, by = 0.01)) |>
  mutate(y = dnorm(x, mean = 1.5, sd = 0.4)) |>
  ggplot() +
  geom_ribbon(
    aes(x = x, ymin = 0.25, ymax = y + 0.25),
    fill = NA,
    color = "black",
    linewidth = 1.5
  ) +
  labs(x = "Songwriting", y = NULL) +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank())
```

:::

:::{.fragment fragment-index=2}

```{r}
#| label: dcm-reliability
#| warning: false
#| out-width: 100%
#| out-height: 100%
#| fig-alt: "Normal distribution with peak at 1.5 on top of categorical x-axis where values less than 0 are labelled 'Not Proficient' and values greater than 0 are labelled 'Proficient.'"

threshold <- 0

tibble(x = seq(-3, 3, by = 0.01)) |>
  mutate(y = dnorm(x, mean = 1.5, sd = 0.4)) |>
  ggplot() +
  geom_ribbon(
    data = ~ filter(., x <= threshold),
    aes(x = x, ymin = 0.25, ymax = y + 0.25),
    color = "black",
    fill = palette_measr[1],
    linewidth = 1.5
  ) +
  geom_ribbon(
    data = ~ filter(., x > threshold),
    aes(x = x, ymin = 0.25, ymax = y + 0.25),
    color = "black",
    fill = palette_measr[3],
    linewidth = 1.5
  ) +
  geom_line(aes(x = x, y = y + 0.25), linewidth = 1.5) +
  geom_rect(
    xmin = -3,
    xmax = threshold,
    ymin = 0,
    ymax = 0.2,
    fill = palette_measr[1]
  ) +
  geom_rect(
    xmin = threshold,
    xmax = 3,
    ymin = 0,
    ymax = 0.2,
    fill = palette_measr[3]
  ) +
  geom_text(
    aes(x = -3 + ((threshold - -3) / 2), y = 0.1, label = "Not Proficient"),
    color = "white"
  ) +
  geom_text(
    aes(x = 3 - ((3 - threshold) / 2), y = 0.1, label = "Proficient"),
    color = "black"
  ) +
  expand_limits(y = 0) +
  labs(x = "Songwriting", y = NULL) +
  theme(
    axis.text.y = element_blank(),
    axis.text.x = element_blank(),
    axis.title.y = element_blank()
  )
```

:::

:::{.fragment fragment-index=4}

```{r}
#| label: dcm-reliability-threshold
#| warning: false
#| out-width: 100%
#| out-height: 100%
#| fig-alt: "Normal distribution with peak at 1.5 on top of categorical x-axis where values less than 1 are labelled 'Not Proficient' and values greater than 1 are labelled 'Proficient.'"

threshold <- 1

tibble(x = seq(-3, 3, by = 0.01)) |>
  mutate(y = dnorm(x, mean = 1.5, sd = 0.4)) |>
  ggplot() +
  geom_ribbon(
    data = ~ filter(., x <= threshold),
    aes(x = x, ymin = 0.25, ymax = y + 0.25),
    color = "black",
    fill = palette_measr[1],
    linewidth = 1.5
  ) +
  geom_ribbon(
    data = ~ filter(., x > threshold),
    aes(x = x, ymin = 0.25, ymax = y + 0.25),
    color = "black",
    fill = palette_measr[3],
    linewidth = 1.5
  ) +
  geom_line(aes(x = x, y = y + 0.25), linewidth = 1.5) +
  geom_rect(
    xmin = -3,
    xmax = threshold,
    ymin = 0,
    ymax = 0.2,
    fill = palette_measr[1]
  ) +
  geom_rect(
    xmin = threshold,
    xmax = 3,
    ymin = 0,
    ymax = 0.2,
    fill = palette_measr[3]
  ) +
  geom_text(
    aes(x = -3 + ((threshold - -3) / 2), y = 0.1, label = "Not Proficient"),
    color = "white"
  ) +
  geom_text(
    aes(x = 3 - ((3 - threshold) / 2), y = 0.1, label = "Proficient"),
    color = "black"
  ) +
  expand_limits(y = 0) +
  labs(x = "Songwriting", y = NULL) +
  theme(
    axis.text.y = element_blank(),
    axis.text.x = element_blank(),
    axis.title.y = element_blank()
  )
```

:::

:::

:::

:::

## When are DCMs appropriate?

Success depends on:

1. Domain definitions
    * What are the attributes we're trying to measure?
    * Are the attributes measurable (e.g., with assessment items)?
  
2. Alignment of purpose between assessment and model
    * Is classification the purpose?

## Example applications

* **Educational measurement:** The competencies that student is or is not proficient in
  * Latent knowledge, skills, or understandings
  * Used for tailored instruction and remediation
  
* **Psychiatric assessment:** The DSM criteria that an individual meets
  * Broader diagnosis of a disorder

## When are DCMs not appropriate?

* When the goal is to place individuals on a scale

* DCMs do not distinguish within classes

:::{.columns}
:::{.column width="50%"}
<br>

```{r}
#| label: red-profiles

taylor_profiles |>
  filter(str_detect(era, "Red")) |>
  select(img, songwriting, production, vocals, ends_with("_color")) |>
  gt() |>
  cols_hide(ends_with("_color")) |>
  cols_label(img = "") |>
  # cols_width(img ~ px(100),
  #            everything() ~ px(150)) |>
  gt_img_rows(columns = img, img_source = "local", height = 75) |>
  fmt_icon(songwriting, fill_color = from_column("s_color"), height = "40px") |>
  fmt_icon(production, fill_color = from_column("p_color"), height = "40px") |>
  fmt_icon(vocals, fill_color = from_column("v_color"), height = "40px") |>
  cols_align("center", -img) |>
  gt_theme_measr() |>
  tab_options(table.font.size = 18)
```

:::

:::{.column width="50%"}
```{r}
#| label: red-scale

table_taylors <- taylors |>
  filter(str_detect(era, "Red"))

prior(normal(0, 1), type = "intercept") |>
  dcmstan::prior_tibble() |>
  parse_dist(prior) |>
  ggplot(aes(xdist = .dist_obj)) +
  stat_slab(color = palette_measr[1], fill = palette_measr[3]) -> base

width <- 0.6
for (i in 1:nrow(table_taylors)) {
  img <- as.raster(image_read(table_taylors$img[i]))

  base <- base +
    annotation_raster(
      img,
      table_taylors$x[i] - (width / 2),
      table_taylors$x[i] + (width / 2),
      table_taylors$y[i],
      table_taylors$y[i] + 0.25
    )
}

base +
  labs(x = "Musical Knowledge", y = NULL) +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank())
```

:::
:::

## Conceptual foundation summary 

:::{.columns}
:::{.column width="50%"}
* DCMs are psychometric models designed to classify
  * We can define our attributes in any way that we choose
  * Items depend on the attribute definitions
  * Classifications are probabilistic
:::

:::{.column .fragment width="50%"}
* DCMs provide valuable information with more feasible data demands than other psychometric models
  * Higher reliability than IRT/MIRT models
  * Complex item structures possible
  * Criterion-referenced interpretations
  * Alignment of assessment goals and psychometric model
:::
:::

# Statistical foundations

## Statistical foundation

* Latent class models use responses to probabilistically place individuals into latent classes

* DCMs are confirmatory latent class models
  * Latent classes specified *a priori* as attribute profiles
  * Q-matrix specifies item-attribute structure
  * Person parameters are attribute proficiency probabilities
  
## Terminology

* **Respondents** (*r*): The individuals from whom behavioral data are collected
  * For today, this is dichotomous assessment item responses
  * Not limited to only item responses in practice

* **Items** (*i*): Assessment questions used to classify/diagnose respondents

* **Attributes** (*a*): Unobserved latent categorical characteristics underlying the behaviors (i.e., diagnostic status)
  * Latent variables

## Attribute profiles

* With binary attributes, there are 2^A^ possible profiles

* Example 3-attribute assessment:

:::{.center .small}
[0, 0, 0]  
[1, 0, 0]  
[0, 1, 0]  
[0, 0, 1]  
[1, 1, 0]  
[1, 0, 1]  
[0, 1, 1]  
[1, 1, 1]
:::

## DCMs as latent class models

$$
\color{#D55E00}{P(X_r=x_r)} = \sum_{c=1}^C\color{#009E73}{\nu_c} \prod_{i=1}^I\color{#56B4E9}{\pi_{ic}^{x_{ir}}(1-\pi_{ic})^{1 - x_{ir}}}
$$

:::{.fragment}
```{=html}
<span class="eqn-box", style="background-color: #D55E00; color: white">Observed data: Probability of observing examinee <em>r</em>'s item reponses</span>
```
:::

:::{.fragment}
```{=html}
<span class="eqn-box", style="background-color: #009E73; color: white">Structural component: Proportion of examinees in each class</span>
```
:::

:::{.fragment}
```{=html}
<span class="eqn-box", style="background-color: #56B4E9; color: white">Measurement component: Product of item response probabilities</span>
```
:::

## Structural models

$$
\color{#D55E00}{P(X_r=x_r)} = \sum_{c=1}^C\color{#009E73}{\nu_c} \prod_{i=1}^I\color{#56B4E9}{\pi_{ic}^{x_{ir}}(1-\pi_{ic})^{1 - x_{ir}}}
$$

```{=html}
<span class="eqn-box", style="background-color: #009E73; color: white; margin-bottom: 0">Structural component: Proportion of examinees in each class</span>
```

* Prevalence of each class in the population
  * &nu;<sub>1</sub> + &nu;<sub>2</sub> + ... + &nu;<sub>c</sub> = 1
* Typically unconstrained
  * Independent attributes ([Lee, 2017](https://mc-stan.org/learn-stan/case-studies/dina_independent.html))
  * Log-linear structural models ([Rupp et al., 2010](https://www.amazon.com/Diagnostic-Measurement-Applications-Methodology-Sciences/dp/1606235273))

## Measurement models

$$
\color{#D55E00}{P(X_r=x_r)} = \sum_{c=1}^C\color{#009E73}{\nu_c} \prod_{i=1}^I\color{#56B4E9}{\pi_{ic}^{x_{ir}}(1-\pi_{ic})^{1 - x_{ir}}}
$$

```{=html}
<span class="eqn-box", style="background-color: #56B4E9; color: white; margin-bottom: 0">Measurement component: Product of item response probabilities</span>
```

* Traditional psychometrics: Item response theory, classical test theory
  * A single, unidimensional construct
  * Student results estimated on a continuum
  * Performance on individual items determined by an "item characteristic curve"
* DCMs: Many different options

## {data-menu-title="First IRT item"}

```{r}
#| label: irt-item
#| out-width: 70%
#| out-height: 100%
#| fig-alt: "A logistic curve showing the probability of providing a correct response."

num_item <- 20

set.seed(121389)

items <- tibble(
  item = seq_len(num_item),
  a = c(1.00, runif(num_item - 1, min = 0.7, max = 3.0)),
  b = c(0.00, rnorm(num_item - 1, mean = 0, sd = 0.6))
)

probs <- expand_grid(theta = seq(-3, 3, by = 0.01), item = seq_len(num_item)) |>
  left_join(items, by = "item") |>
  mutate(
    log_odds = a * (theta - b),
    prob_1 = 1 / (1 + exp(-1 * log_odds)),
    prob_0 = 1 - prob_1
  )

ggplot(probs, aes(x = theta)) +
  geom_line(
    data = ~ filter(.x, item == 1),
    aes(y = prob_1),
    color = palette_okabeito[1]
  ) +
  expand_limits(x = c(-3, 3), y = c(0, 1)) +
  scale_x_continuous(breaks = seq(-3, 3, by = 1)) +
  scale_y_percent(breaks = seq(0, 1, by = 0.2)) +
  labs(
    x = "Knowledge, Skills, and Understandings",
    y = "Probability of Response"
  )
```

## {data-menu-title="Second IRT item"}

```{r}
#| label: irt-item2
#| out-width: 70%
#| out-height: 100%
#| fig-alt: "Two logistic curves showing the probability of providing a correct response for two items."

ggplot(probs, aes(x = theta)) +
  geom_line(
    data = ~ filter(.x, item == 1),
    aes(y = prob_1),
    color = palette_okabeito[1]
  ) +
  geom_line(
    data = ~ filter(.x, item == 2),
    aes(y = prob_1),
    color = palette_okabeito[2]
  ) +
  expand_limits(x = c(-3, 3), y = c(0, 1)) +
  scale_x_continuous(breaks = seq(-3, 3, by = 1)) +
  scale_y_percent(breaks = seq(0, 1, by = 0.2)) +
  labs(
    x = "Knowledge, Skills, and Understandings",
    y = "Probability of Response"
  )
```

## {data-menu-title="Third IRT item"}

```{r}
#| label: irt-item3
#| out-width: 70%
#| out-height: 100%
#| fig-alt: "Three logistic curves showing the probability of providing a correct response for three items."

ggplot(probs, aes(x = theta)) +
  geom_line(
    data = ~ filter(.x, item == 1),
    aes(y = prob_1),
    color = palette_okabeito[1]
  ) +
  geom_line(
    data = ~ filter(.x, item == 2),
    aes(y = prob_1),
    color = palette_okabeito[2]
  ) +
  geom_line(
    data = ~ filter(.x, item == 3),
    aes(y = prob_1),
    color = palette_okabeito[3]
  ) +
  expand_limits(x = c(-3, 3), y = c(0, 1)) +
  scale_x_continuous(breaks = seq(-3, 3, by = 1)) +
  scale_y_percent(breaks = seq(0, 1, by = 0.2)) +
  labs(
    x = "Knowledge, Skills, and Understandings",
    y = "Probability of Response"
  )
```

## {data-menu-title="Incorrect IRT item"}

```{r}
#| label: irt-item4
#| out-width: 70%
#| out-height: 100%
#| fig-alt: "Three logistic curves showing the probability of providing a correct response for three items, and 1 logistic curve showing the probabiliyt of providing an incorrect response for a fourth item."

ggplot(probs, aes(x = theta)) +
  geom_line(
    data = ~ filter(.x, item == 1),
    aes(y = prob_1),
    color = palette_okabeito[1]
  ) +
  geom_line(
    data = ~ filter(.x, item == 2),
    aes(y = prob_1),
    color = palette_okabeito[2]
  ) +
  geom_line(
    data = ~ filter(.x, item == 3),
    aes(y = prob_1),
    color = palette_okabeito[3]
  ) +
  geom_line(
    data = ~ filter(.x, item == 4),
    aes(y = prob_0),
    color = palette_okabeito[5]
  ) +
  expand_limits(x = c(-3, 3), y = c(0, 1)) +
  scale_x_continuous(breaks = seq(-3, 3, by = 1)) +
  scale_y_percent(breaks = seq(0, 1, by = 0.2)) +
  labs(
    x = "Knowledge, Skills, and Understandings",
    y = "Probability of Response"
  )
```


## Diagnostic assessment items

* Can be multidimensional

* No continuum of student achievement

* Categorical constructs
  * Usually binary (e.g., master/nonmaster, proficient/not proficient)
  
## DCM measurement models

* Items can measure one or both attributes

* Different DCMs define &pi;<sub>ic</sub> in different ways
  * Each DCM makes different assumptions about how attributes proficiencies combine/interact to produce an item response

* Item characteristic bar charts

## Loglinear cognitive diagnostic model (LCDM)

:::{.columns}
:::{.column width="30%"}
* [Henson et al. (2009)](https://doi.org/10.1007/s11336-008-9089-5)

* Different response probabilities for each class (partially compensatory)

* This will be our focus
:::

:::{.column width="70%"}
```{r}
#| label: lcdm-icb
#| out-width: 100%
#| out-height: 50%
#| fig-alt: "Bar graph showing a high probability of providing a correct response when proficient on both attribute 1 and attribute 2 and a moderate probability when only proficient on one of the attributes."

icb <- tribble(
  ~profile    , ~single , ~comp , ~noncomp , ~partial , ~lcdm ,
  "[0, 0, 0]" , 0.15    , 0.15  , 0.15     , 0.15     , 0.15  ,
  "[1, 0, 0]" , 0.90    , 0.90  , 0.15     , 0.50     , 0.60  ,
  "[0, 1, 0]" , 0.15    , 0.90  , 0.15     , 0.35     , 0.40  ,
  "[0, 0, 1]" , 0.15    , 0.15  , 0.15     , 0.15     , 0.15  ,
  "[1, 1, 0]" , 0.90    , 0.90  , 0.90     , 0.75     , 0.90  ,
  "[1, 0, 1]" , 0.90    , 0.90  , 0.15     , 0.50     , 0.60  ,
  "[0, 1, 1]" , 0.15    , 0.90  , 0.15     , 0.35     , 0.40  ,
  "[1, 1, 1]" , 0.90    , 0.90  , 0.90     , 0.75     , 0.90
) |>
  mutate(profile = fct_inorder(profile))

icb |>
  mutate(
    prof = case_when(
      lcdm < 0.2 ~ "FALSE",
      lcdm > 0.8 ~ "TRUE",
      TRUE ~ "Partial"
    )
  ) |>
  ggplot(aes(x = profile, y = lcdm, fill = prof)) +
  geom_col(show.legend = FALSE) +
  expand_limits(y = c(0, 1)) +
  scale_y_percent(breaks = seq(0, 1, by = 0.2)) +
  scale_fill_manual(
    values = c(
      "TRUE" = palette_measr[4],
      "FALSE" = palette_measr[2],
      "Partial" = "#4B3F72"
    )
  ) +
  labs(x = "Profile", y = "Probability of Correct Response")
```

:::
:::

## Simple structure LCDM

Item measures only 1 attribute

$$
\text{logit}(X_i = 1) = \color{#D7263D}{\lambda_{i,0}} + \color{#219EBC}{\lambda_{i,1(1)}}\color{#009E73}{\alpha}
$$

:::{.fragment}
```{=html}
<span class="eqn-box2", style="background-color: #D7263D; color: white">&lambda;<sub>i,0</sub>: Log-odds when not proficient</span>
```
:::

:::{.fragment}
```{=html}
<span class="eqn-box2", style="background-color: #219EBC; color: white">&lambda;<sub>i,1(1)</sub>: Increase in log-odds when proficient</span>
```
:::

:::{.fragment}
```{=html}
<span class="eqn-box2", style="background-color: #009E73; color: white">&alpha;: Attribute proficiency status (either 0 or 1)</span>
```
:::

## Subscript notation

:::{.columns}
:::{.column .center .larger width="40%"}

<br><br>

```{=html}
&lambda;<sub>i,e(&alpha;<sub>1</sub>)</sub>
```

:::

:::{.column width="60%"}
:::{.fragment}
- *i* = The item to which the parameter belongs
:::

:::{.fragment}
- *e* = The level of the effect
  - 0 = intercept
  - 1 = main effect
  - 2 = two-way interaction
  - 3 = three-way interaction
  - Etc.
:::

:::{.fragment}
- (&alpha;<sub>1</sub>,...) = The attributes to which the effect applies
  - The same number of attributes as listed in subscript 2
:::
:::
:::

## Complex structure LCDM

Item measures multiple attributes

$$
\text{logit}(X_i = 1) = \color{#D7263D}{\lambda_{i,0}} + \color{#4B3F72}{\lambda_{i,1(1)}\alpha_1} + \color{#9589BE}{\lambda_{i,1(2)}\alpha_2} +
\color{#219EBC}{\lambda_{i,2(1,2)}\alpha_1\alpha_2}
$$

:::{.fragment}
```{=html}
<span class="eqn-box2", style="background-color: #D7263D; color: white">Log-odds when proficient in neither attribute</span>
```
:::

:::{.fragment}
```{=html}
<span class="eqn-box2", style="background-color: #4B3F72; color: white">Increase in log-odds when proficient in attribute 1</span>
```
:::

:::{.fragment}
```{=html}
<span class="eqn-box2", style="background-color: #9589BE; color: white">Increase in log-odds when proficient in attribute 2</span>
```
:::

:::{.fragment}
```{=html}
<span class="eqn-box2", style="background-color: #219EBC; color: white">Change in log-odds when proficient in both attributes</span>
```
:::


## Defining DCM structures

* Attribute and item relationships are defined in the Q-matrix

* Q-matrix
  * *I* $\times$ *A* matrix
  * 0 = Attribute is not measured by the item
  * 1 = Attribute is measured by the item
  
## The LCDM as a general DCM

* So called "general" DCM because the LCDM subsumes other DCMs

* Constraints on item parameters make LCDM equivalent to other DCMs (e.g., DINA and DINO)

## From model parameters to respondents

* Respondent estimates come from combining the estimated model parameters with the response data

* For DCMs, a similar process to that for IRT

## IRT respondent estimates

:::{.columns}

:::{.column width="30%"}

* Multiply the ICCs together
  * Multiply the response probabilities together for each value of the trait

* Student estimate is the peak of the curve

* Spread of the curve represents uncertainty in estimate

:::

:::{.column width="70%"}

:::{.r-stack}

:::{.fragment}

```{r}
#| label: irt-item4
#| out-width: "100%"
#| out-height: "100%"
```

:::

:::{.fragment}

```{r}
#| label: irt-student-estimate
#| out-width: 100%
#| out-height: 100%
#| fig-alt: "Line graph in the shape of normal distribution. A dashed vertical line indicates the location of the peak of the curve."

probs |>
  filter(item <= 4) |>
  mutate(
    response_prob = case_when(item %in% c(4) ~ prob_0, .default = prob_1)
  ) |>
  summarize(logll = sum(log(response_prob)), .by = theta) |>
  mutate(likelihood = exp(logll)) |>
  ggplot(aes(x = theta)) +
  geom_line(aes(y = likelihood)) +
  geom_vline(
    aes(xintercept = theta[which.max(likelihood)]),
    linetype = "dashed"
  ) +
  scale_x_continuous(breaks = seq(-3, 3, by = 1)) +
  labs(x = "Knowledge, Skills, and Understandings", y = "Likelihood")
```

:::

:::

:::

:::

## DCM respondent estimates

:::{.columns}

:::{.column width="30%"}

:::{.fragment fragment-index=1}

* Multiply the response probabilities together for each class

:::

:::{.fragment fragment-index=3}

* Multiply the item response likelihoods by structural parameters

:::

:::{.fragment fragment-index=5}

* Class probabilities are the class likelihoods divided by the total likelihood

:::

:::

:::{.column width="70%"}

:::{.r-stack}

```{r}
#| label: dcm-icbs
#| echo: false
#| message: false
#| out-width: 100%
#| out-height: 100%
#| fig-alt: "Bar graphs showing the response probabilities for each class for 4 items, where the fourth item was answered incorrectly."

library(posterior)
taylor_data <- read_rds(here("materials", "slides", "data", "taylor-data.rds"))
taylor_qmatrix <- read_rds(here(
  "materials",
  "slides",
  "data",
  "taylor-qmatrix.rds"
))

taylor_spec <- dcm_specify(
  qmatrix = taylor_qmatrix,
  measurement_model = lcdm(),
  structural_model = unconstrained()
)

taylor_dcm <- dcm_estimate(
  taylor_spec,
  data = taylor_data,
  identifier = "album",
  method = "mcmc",
  backend = "cmdstanr",
  chains = 4,
  parallel_chains = 4,
  seed = 121389,
  iter_warmup = 1000,
  iter_sampling = 500,
  file = here("materials", "slides", "fits", "taylor-lcdm-uncst")
)

taylor_pi <- taylor_dcm@model |>
  as_draws_df() |>
  subset_draws(variable = "pi") |>
  as_tibble() |>
  pivot_longer(cols = -c(.chain, .iteration, .draw)) |>
  summarize(value = mean(value), .by = name) |>
  mutate(
    item = str_replace_all(name, "pi\\[([0-9]*),([0-9]*)\\]", "\\1"),
    class = str_replace_all(name, "pi\\[([0-9]*),([0-9]*)\\]", "\\2")
  ) |>
  select(item, class, pi = value) |>
  mutate(across(where(is.character), as.integer)) |>
  left_join(
    measr_extract(taylor_dcm, "classes") |>
      rename(label = class) |>
      rowid_to_column(var = "class"),
    join_by(class)
  ) |>
  mutate(label = str_replace_all(label, ",", ", "), label = fct_inorder(label))

taylor_pi |>
  filter(item <= 4) |>
  mutate(pi = case_when(item == 4 ~ 1 - pi, .default = pi)) |>
  ggplot(aes(x = label, y = pi)) +
  facet_wrap(vars(item), ncol = 1) +
  geom_col(aes(fill = factor(item)), show.legend = FALSE) +
  expand_limits(y = c(0, 1)) +
  scale_y_percent() +
  scale_fill_manual(values = c(palette_okabeito[c(1:3, 5)])) +
  labs(x = "Profile", y = "Probability of Response") +
  theme(strip.text.x = element_blank())
```

:::{.fragment fragment-index=2}

```{r}
#| label: dcm-class-item-likelihood
#| out-width: 100%
#| out-height: 100%
#| fig-alt: "Bar graph showing the product of the item response probabilities for each class."

taylor_pi |>
  filter(item <= 4) |>
  mutate(pi = case_when(item == 4 ~ 1 - pi, .default = pi)) |>
  summarize(pi_prod = prod(pi), .by = label) |>
  ggplot(aes(x = label, y = pi_prod)) +
  geom_col(fill = "black", show.legend = FALSE) +
  scale_y_continuous(expand = c(0.01, 0), labels = \(x) sprintf("%0.3f", x)) +
  labs(x = "Profile", y = "Likelihood of Responses") +
  theme(strip.text.x = element_blank())
```

:::

:::{.fragment fragment-index=4}

```{r}
#| label: dcm-class-likelihood
#| out-width: 100%
#| out-height: 100%
#| fig-alt: "Bar graph showing the likelihood for each class."

strc <- taylor_dcm@model |>
  as_draws_df() |>
  subset_draws(variable = "Vc") |>
  as_tibble() |>
  pivot_longer(cols = -c(.chain, .iteration, .draw)) |>
  summarize(value = mean(value), .by = name) |>
  mutate(
    name = str_replace_all(name, "Vc\\[([0-9]*)\\]", "\\1"),
    name = as.integer(name)
  )

taylor_pi |>
  filter(item <= 4) |>
  mutate(pi = case_when(item == 4 ~ 1 - pi, .default = pi)) |>
  summarize(pi_prod = prod(pi), .by = c(class, label)) |>
  left_join(strc, join_by(class == name)) |>
  mutate(likelihood = pi_prod * value) |>
  ggplot(aes(x = label, y = likelihood)) +
  geom_col(fill = "black", show.legend = FALSE) +
  scale_y_continuous(expand = c(0.01, 0), labels = \(x) sprintf("%0.3f", x)) +
  labs(x = "Profile", y = "Likelihood") +
  theme(strip.text.x = element_blank())
```

:::

:::{.fragment fragment-index=6}

```{r}
#| label: dcm-class-probabilities
#| out-width: 100%
#| out-height: 100%
#| fig-alt: "Bar graph showing the probability that the respondent belongs to each class."

class_probs <- taylor_pi |>
  filter(item <= 4) |>
  mutate(pi = case_when(item == 4 ~ 1 - pi, .default = pi)) |>
  summarize(pi_prod = prod(pi), .by = c(class, label)) |>
  left_join(strc, join_by(class == name)) |>
  mutate(likelihood = pi_prod * value, prob = likelihood / sum(likelihood))

class_probs |>
  ggplot(aes(x = label, y = prob)) +
  geom_col(fill = "black", show.legend = FALSE) +
  scale_y_percent() +
  labs(x = "Profile", y = "Class Probability") +
  theme(strip.text.x = element_blank())
```

:::

:::

:::

:::

## From class to attribute probabilities

:::{.columns}

:::{.column width="40%"}

* For each attribute, sum the class probabilities where that attribute is present

:::{.fragment fragment-index=1}
Songwriting: 84.2%
:::

:::{.fragment fragment-index=2}
Production: 45.4%
:::

:::{.fragment fragment-index=3}
Vocals: 88.4%
:::

:::

:::{.column width="60%"}

:::{.r-stack}

:::{.fragment fragment-index=1}

```{r}
#| label: songwriting-prob

att_probs <- create_profiles(3) |>
  rowid_to_column(var = "class") |>
  left_join(class_probs, join_by(class)) |>
  select(
    songwriting = att1,
    production = att2,
    vocals = att3,
    probability = prob
  )

att_probs |>
  add_row(
    probability = sum(att_probs$probability[which(att_probs$songwriting == 1)])
  ) |>
  gt() |>
  sub_missing(missing_text = "") |>
  fmt_number(columns = probability, decimals = 3) |>
  cols_align("center", everything()) |>
  gt_theme_measr() |>
  tab_style(
    style = list(cell_fill(color = palette_measr[3])),
    locations = cells_body(columns = songwriting, rows = 1:8)
  ) |>
  tab_style(
    style = list(
      cell_fill(color = palette_measr[1]),
      cell_text(color = "white")
    ),
    locations = cells_body(
      columns = everything(),
      rows = which(att_probs$songwriting == 1)
    )
  ) |>
  tab_style(
    style = list(cell_text(color = palette_measr[2], weight = "bold")),
    locations = cells_body(columns = probability, rows = 9)
  ) |>
  tab_options(table.font.size = 24)
```

:::

:::{.fragment fragment-index=2}

```{r}
#| label: production-prob

att_probs |>
  add_row(
    probability = sum(att_probs$probability[which(att_probs$production == 1)])
  ) |>
  gt() |>
  sub_missing(missing_text = "") |>
  fmt_number(columns = probability, decimals = 3) |>
  cols_align("center", everything()) |>
  gt_theme_measr() |>
  tab_style(
    style = list(cell_fill(color = palette_measr[3])),
    locations = cells_body(columns = production, rows = 1:8)
  ) |>
  tab_style(
    style = list(
      cell_fill(color = palette_measr[1]),
      cell_text(color = "white")
    ),
    locations = cells_body(
      columns = everything(),
      rows = which(att_probs$production == 1)
    )
  ) |>
  tab_style(
    style = list(cell_text(color = palette_measr[2], weight = "bold")),
    locations = cells_body(columns = probability, rows = 9)
  ) |>
  tab_options(table.font.size = 24)
```

:::

:::{.fragment fragment-index=3}

```{r}
#| lable: vocal-prob

att_probs |>
  add_row(
    probability = sum(att_probs$probability[which(att_probs$vocals == 1)])
  ) |>
  gt() |>
  sub_missing(missing_text = "") |>
  fmt_number(columns = probability, decimals = 3) |>
  cols_align("center", everything()) |>
  gt_theme_measr() |>
  tab_style(
    style = list(cell_fill(color = palette_measr[3])),
    locations = cells_body(columns = vocals, rows = 1:8)
  ) |>
  tab_style(
    style = list(
      cell_fill(color = palette_measr[1]),
      cell_text(color = "white")
    ),
    locations = cells_body(
      columns = everything(),
      rows = which(att_probs$vocals == 1)
    )
  ) |>
  tab_style(
    style = list(cell_text(color = palette_measr[2], weight = "bold")),
    locations = cells_body(columns = probability, rows = 9)
  ) |>
  tab_options(table.font.size = 24)
```

:::

:::

:::

:::

## The rest of today

* Estimating DCMs with [Stan](https://mc-stan.org) and [measr](https://measr.r-dcm.org)

* Evaluating DCMs with [measr](https://measr.r-dcm.org)

# {.closing data-menu-title="Closing"}

<br>

:::{.end-title color="white" font-size="200%"}
Diagnostic classification models
:::

:::{.end-subtitle}
A brief introduction
:::

:::{.center}
<https://learn.r-dcm.org>
:::
