---
title: Evaluating diagnostic <br> classification models
subtitle: With Stan and measr
author: W. Jake Thompson, Ph.D.
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(ggmeasr)
library(ggdist)
library(magick)
library(distributional)
library(knitr)
library(measr)
library(here)
library(fs)
library(downlit)
library(cmdstanr)
library(countdown)
library(posterior)

set_theme(plot_margin = margin(5, 0, 0, 0))

# copy grid-worms
if (!dir_exists(here("_site/materials/slides/"))) {
  dir_create(here("_site/materials/slides/"))
}
dir_copy(
  here("materials", "slides", "grid-worms"),
  here("_site/materials/slides/grid-worms"),
  overwrite = TRUE
)
```

## Model evaluation

* **Absolute fit**: How well does the model represent the observed data?
  * Model-level
  * Item-level

* **Relative fit**: How do multiple models compare to each other?

* **Reliability**: How consistent and accurate are the classifications?

# Absolute model fit

## Model-level absolute fit

* Limited information indices based on parameter point estimates
  * M~2~ statistic ([Liu et al., 2016](https://doi.org/10.3102/1076998615621293))
  
* Posterior predictive model checks (PPMCs)

## Calculating limited information indices

```{r}
#| label: read-ecpe

library(dcmdata)

ecpe_spec <- dcm_specify(
  qmatrix = ecpe_qmatrix, # <1>
  identifier = "item_id", # <1>
  measurement_model = lcdm(), # <2>
  structural_model = unconstrained() # <2>
)

ecpe_lcdm <- dcm_estimate(
  dcm_spec = ecpe_spec, # <1>
  data = ecpe_data, # <2>
  identifier = "resp_id", # <2>
  method = "pathfinder", # <3>
  backend = "cmdstanr", # <3>
  single_path_draws = 1000, # <4>
  num_paths = 10, # <4>
  draws = 2000, # <4>
  file = here("materials", "slides", "fits", "ecpe-lcdm-uncst") # <5>
)

ecpe_lcdm <- add_fit(
  ecpe_lcdm,
  method = c("m2", "ppmc"),
  model_fit = "raw_score",
  item_fit = c("pvalue", "conditional_prob", "odds_ratio")
)
ecpe_lcdm <- add_criterion(ecpe_lcdm, criterion = c("waic", "loo"))
ecpe_lcdm <- add_reliability(ecpe_lcdm)
```

* Estimates model fit using univariate and bivariate item relationships

* *p*-values less than .05 indicate poor fit

```{r}
#| label: ecpe-m2
#| echo: true

fit_m2(ecpe_lcdm)
```

* Also includes the root mean square error of approximation (RMSEA) and the standardized root mean square residual (SRMSR)

## Exercise 1 {.exercise}

* Open `evaluation.Rmd` and run the `setup` chunk.

* Calculate the M~2~ statistic for the ROAR-PA LCDM model

* Does the model fit the data?

{{< countdown "3:00" bottom="15%" >}}

## {.exercise data-menu-title="Solution 1"}

```{r}
#| label: read-roarpa

roarpa_data <- roarpa_data |>
  filter(complete.cases(roarpa_data))

roarpa_spec <- dcm_specify(
  qmatrix = roarpa_qmatrix,
  identifier = "item",
  measurement_model = lcdm(),
  structural_model = unconstrained()
)

roarpa_lcdm <- dcm_estimate(
  roarpa_spec,
  data = roarpa_data,
  identifier = "id",
  method = "variational",
  backend = "rstan",
  seed = 19891213,
  file = here("materials", "slides", "fits", "roarpa-lcdm-uncst")
)

roarpa_lcdm <- add_fit(
  roarpa_lcdm,
  method = c("m2", "ppmc"),
  model_fit = "raw_score",
  item_fit = c("pvalue", "conditional_prob", "odds_ratio")
)
roarpa_lcdm <- add_criterion(roarpa_lcdm, criterion = c("waic", "loo"))
roarpa_lcdm <- add_reliability(roarpa_lcdm)
```

```{r}
#| label: roarpa-m2
#| echo: true

fit_m2(roarpa_lcdm)
```

* The null hypothesis is that our model fits the data
  * *p*-values > .05 indicate acceptable model fit

## PPMC: Raw score distribution

:::{.columns}

:::{.column width="30%"}
* For each iteration, calculate the total number of respondents at each score point

:::{.fragment fragment-index=2}

* Calculate the expected number of respondents at each score point

:::

:::{.fragment fragment-index=3}

* Calculate the <span style="color: #D7263D;">observed</span> number of respondents at each score point

:::

:::

:::{.column width="70%"}

:::{.r-stack}

:::{.fragment fragment-index=1}

```{r}
#| label: score-dist
#| out-width: 100%
#| out-height: 50%
#| fig-alt: |
#|   Scatter plot showing the number of respondents at each score point in each
#|   iteration.

draws <- fit_ppmc(ecpe_lcdm, model_fit = "raw_score", return_draws = 2000)

p <- draws$ppmc_raw_score$rawscore_samples |>
  pluck("rawscores") |>
  unnest(raw_scores) |>
  slice_sample(n = 500, by = raw_score) |>
  ggplot() +
  geom_point(
    aes(x = factor(raw_score), y = n),
    position = position_jitter(height = 0, seed = 1213),
    alpha = 0.2,
    color = palette_measr[4]
  ) +
  scale_y_comma() +
  labs(x = "Correct Responses", y = "Respondents")

p
```

:::

:::{.fragment fragment-index=2}

```{r}
#| label: exp-score
#| out-width: 100%
#| out-height: 50%
#| fig-alt: |
#|   Scatter plot showing the number of respondents at each score point in each
#|   iteration with the average number of respondents overlayed.

exp_scores <- draws$ppmc_raw_score$rawscore_samples |>
  pluck("rawscores") |>
  unnest(raw_scores) |>
  summarize(n = mean(n), .by = raw_score)

p <- p +
  geom_point(
    data = exp_scores,
    aes(x = factor(raw_score), y = n),
    color = palette_measr[1],
    shape = 18,
    size = 5
  ) +
  geom_line(
    data = exp_scores,
    aes(x = factor(raw_score), y = n),
    group = 1,
    color = palette_measr[1]
  )

p
```

:::

:::{.fragment fragment-index=3}

```{r}
#| label: obs-score
#| out-width: 100%
#| out-height: 50%
#| fig-alt: |
#|   Scatter plot showing the number of respondents at each score point in each
#|   iteration with the average and observed number of respondents overlayed.

obs_scores <- ecpe_data |>
  pivot_longer(-resp_id) |>
  summarize(score = sum(value), .by = resp_id) |>
  count(score)

p <- p +
  geom_point(
    data = obs_scores,
    aes(x = factor(score), y = n),
    color = palette_measr[2],
    shape = 16,
    size = 5
  ) +
  geom_line(
    data = obs_scores,
    aes(x = factor(score), y = n),
    color = palette_measr[2],
    group = 1
  )

p
```

:::

:::

:::

:::

## PPMC: $\chi^2$

:::{.columns}

:::{.column width="30%"}
* Calculate a $\chi^2$-like statistic comparing the number of respondents at each score point in each iteration to the expectation

:::{.fragment fragment-index=2}

* Calculate the $\chi^2$ value comparing the <span style="color: #D7263D;">observed</span> data to the expectation

:::

:::

:::{.column width="70%"}

:::{.r-stack}

:::{.fragment fragment-index=1}

```{r}
#| label: chisq-dist
#| warning: false
#| out-width: 100%
#| out-height: 50%
#| fig-alt: "Histogram of the chi-square values from each iteration."

chisq <- draws$ppmc_raw_score$chisq_samples$chisqs

p <- ggplot() +
  geom_histogram(
    aes(x = chisq),
    binwidth = 2,
    boundary = 0,
    fill = palette_measr[1],
    color = palette_measr[4]
  ) +
  scale_x_continuous(limits = c(0, 75)) +
  labs(x = "&chi;<sup>2</sup><sub>rep</sub>", y = "Replications") +
  theme(axis.title.x = ggtext::element_markdown(family = "sans"))

p
```

:::

:::{.fragment fragment-index=2}

```{r}
#| label: chisq-obs
#| warning: false
#| out-width: 100%
#| out-height: 50%
#| fig-alt: |
#|   Histogram of the chi-square values from each iteration with a dashed
#|   vertical line indicating the value from the observed data.

obs_chisq <- draws$ppmc_raw_score$obs_chisq

p <- p +
  geom_vline(
    xintercept = obs_chisq,
    linetype = "dashed",
    color = palette_measr[2]
  ) +
  scale_x_continuous(limits = c(0, 1000))

p
```

:::

:::

:::

:::

## PPMC: *ppp*

* Calculate the proportion of iterations where the $\chi^2$-like statistic from replicated data set exceeds the observed data statistic
  * Posterior predictive *p*-value (*ppp*)
  
* Very high values (e.g., >.975) or very low values (e.g., <.025) indicate our observed value is far in the tails of what the model would predict
  * Poor model fit

* Ideally, we'd like *ppp* values close to .5

## PPMCs with measr

```{r}
#| label: ecpe-raw-score
#| echo: true

fit_ppmc(ecpe_lcdm, model_fit = "raw_score")
```

## Exercise 2 {.exercise}

* Calculate the raw score PPMC for the ROAR-PA LCDM

* Does the model fit the observed data?

{{< countdown "4:00" bottom="15%" >}}

## {.exercise data-menu-title="Solution 2"}

```{r}
#| label: roarpa-raw-score
#| echo: true

fit_ppmc(roarpa_lcdm, model_fit = "raw_score")
```

* The *ppp* value is <.001
* Model fit appears to be poor.
* Why the difference from the M~2~ conclusion?

## ROAR-PA raw score distribution {.exercise}

:::{.columns}

:::{.column width="30%"}
* Model is underestimating the number of students with the highest scores
* M~2~ is *limited* information---we may miss aspects of the data that come from higher order interactions between items

:::

:::{.column width="70%"}
```{r}
#| label: roarpa-raw-score-plot
#| warning: false
#| out-width: 100%
#| out-height: 50%
#| fig-alt: |
#|   Scatter plot showing the number of respondents at each score point in each
#|   iteration with the average and observed number of respondents overlayed.

draws <- fit_ppmc(roarpa_lcdm, model_fit = "raw_score", return_draws = 2000)

exp_scores <- draws$ppmc_raw_score$rawscore_samples |>
  pluck("rawscores") |>
  unnest(raw_scores) |>
  summarize(n = mean(n), .by = raw_score)

obs_scores <- roarpa_data |>
  filter(complete.cases(roarpa_data)) |>
  pivot_longer(-id) |>
  summarize(score = sum(value), .by = id) |>
  count(score)

draws$ppmc_raw_score$rawscore_samples |>
  pluck("rawscores") |>
  unnest(raw_scores) |>
  slice_sample(n = 500, by = raw_score) |>
  ggplot() +
  geom_point(
    aes(x = factor(raw_score), y = n),
    position = position_jitter(height = 0, seed = 1213),
    alpha = 0.2,
    color = palette_measr[4]
  ) +
  geom_point(
    data = exp_scores,
    aes(x = factor(raw_score), y = n),
    color = palette_measr[1],
    shape = 18,
    size = 5
  ) +
  geom_line(
    data = exp_scores,
    aes(x = factor(raw_score), y = n),
    group = 1,
    color = palette_measr[1]
  ) +
  geom_point(
    data = obs_scores,
    aes(x = factor(score), y = n),
    color = palette_measr[2],
    shape = 16,
    size = 5
  ) +
  geom_line(
    data = obs_scores,
    aes(x = factor(score), y = n),
    color = palette_measr[2],
    group = 1
  ) +
  scale_y_comma() +
  scale_x_discrete(
    limits = factor(13:57),
    breaks = factor(seq(13, 57, by = 2))
  ) +
  labs(x = "Correct Responses", y = "Respondents")
```

:::
:::


## Item-level fit

* Diagnose problems with model-level

* Identify particular items that may not be performing as expected

* Identify potential dimensionality issues

## Item-level fit with measr

* Currently support three measures of item-level fit using PPMCs:
  * Overall item *p*-values
  * Conditional probability of each class providing a correct response ($\pi$ matrix)
  * Item pair odds ratios

## Calculating item-level fit

```{r}
#| label: ecpe-odds-ratio
#| echo: true

fit_ppmc(ecpe_lcdm, item_fit = "odds_ratio")
```


## Extracting item-level fit

```{r}
#| label: ecpe-pvalue
#| echo: true

measr_extract(ecpe_lcdm, "ppmc_pvalue")
```

## Flagging item-level fit

:::{.panel-tabset}

### *p*-values

```{r}
#| label: ecpe-pvalue-flags
#| echo: true

measr_extract(ecpe_lcdm, "ppmc_pvalue_flags")
```

### Conditional probabilities

:::{.new-height style="--code-height: 460px;"}
```{r}
#| label: ecpe-cond-prob-flags
#| echo: true

measr_extract(ecpe_lcdm, "ppmc_conditional_prob_flags")
```

:::

### Odds ratios

:::{.new-height style="--code-height: 460px;"}
```{r}
#| label: ecpe-odds-ratio-flags
#| echo: true

measr_extract(ecpe_lcdm, "ppmc_odds_ratio_flags")
```

:::
:::

## Patterns of item-level misfit

```{r}
#| label: flag-patterns
#| echo: true
#| code-line-numbers: "|8-9,11"

measr_extract(ecpe_lcdm, "ppmc_conditional_prob_flags") |>
  count(class, name = "flags") |>
  left_join(measr_extract(ecpe_lcdm, "strc_param"), by = join_by(class)) |>
  arrange(desc(flags))
```

## Exercise 3 {.exercise}

* Calculate PPMCs for the conditional probabilities and odds ratios for the ROAR-PA model

* What do the results tell us about the model?

{{< countdown "5:00" bottom="15%" >}}

## {.exercise data-menu-title="Solution 3"}

:::{.panel-tabset}

### Conditional probabilities

```{r}
#| label: roarpa-cond-prob
#| echo: true

measr_extract(roarpa_lcdm, "ppmc_conditional_prob_flags") |>
  count(class, name = "flags") |>
  left_join(measr_extract(roarpa_lcdm, "strc_param"), by = join_by(class)) |>
  arrange(desc(flags))
```

### Odds ratios

```{r}
#| label: roarpa-odds-ratio
#| echo: true

measr_extract(roarpa_lcdm, "ppmc_odds_ratio_flags")
```

:::

# Relative model fit

## Model comparisons 

* Doesn't give us information whether or not a model fits the data, only compares competing models to each other
  * Should be evaluated in conjunction with absolute model fit

* Implemented with the [loo](https://mc-stan.org/loo) package
  * PSIS-LOO ([Vehtari, 2017](https://doi.org/10.1007/s11222-016-9696-4))
  * WAIC ([Watanabe, 2010](https://jmlr.org/papers/v11/watanabe10a.html))

## Relative fit with measr

```{r}
#| label: ecpe-loo
#| echo: true

loo(ecpe_lcdm)

waic(ecpe_lcdm)
```

## Comparing models

First, we need another model to compare

:::{.new-height style="--code-height: 480px;"}
```{r}
#| label: ecpe-dina
#| echo: true
#| code-line-numbers: "|4,9"

ecpe_dina_spec <- dcm_specify(
  qmatrix = ecpe_qmatrix,
  identifier = "item_id",
  measurement_model = dina(),
  structural_model = unconstrained()
)

ecpe_dina <- dcm_estimate(
  ecpe_dina_spec,
  data = ecpe_data,
  identifier = "resp_id",
  method = "pathfinder",
  backend = "cmdstanr",
  single_path_draws = 1000,
  num_paths = 10,
  draws = 2000,
  file = here("materials", "slides", "fits", "ecpe-dina-uncst")
)
```

:::

## `loo_compare()`

```{r}
#| label: ecpe-dina-loo
#| include: false

ecpe_dina <- add_criterion(ecpe_dina, criterion = c("loo", "waic"))
```

```{r}
#| label: ecpe-compare-loo
#| echo: true

loo_compare(ecpe_lcdm, ecpe_dina)
```

* LCDM is the preferred model
  * Preferred does not imply "good"
  * Remember, the LCDM showed poor absolute fit

* Difference is much larger than the standard error of the difference
  * Approximate threshold: Absolute difference is greater than 2.5 &times; standard error of the difference ([Bengio & Grandvalet, 2004](https://www.jmlr.org/papers/v5/grandvalet04a.html))

## Exercise 4 {.exercise}

* Estimate a DINA model for the ROAR-PA data

* Use `loo_compare()` to compare the LCDM and DINA models
  * What do the findings tell us?
  * Can you explain the results?

{{< countdown "15:00" bottom="15%" >}}

## {.exercise data-menu-title="Solution 4"}

:::{.panel-tabset}

### Estimate DINA model

```{r}
#| label: roarpa-dina
#| echo: true

roarpa_dina_spec <- dcm_specify(
  qmatrix = roarpa_qmatrix,
  identifier = "item",
  measurement_model = dina(),
  structural_model = unconstrained()
)

roarpa_dina <- dcm_estimate(
  roarpa_dina_spec,
  data = roarpa_data,
  identifier = "id",
  method = "variational",
  backend = "rstan",
  seed = 19891213,
  file = here("materials", "slides", "fits", "roarpa-dina-uncst")
)
```

### Compare models

```{r}
#| label: roarpa-dina-criteria
#| incldue: false

roarpa_dina <- add_criterion(roarpa_dina, criterion = c("loo", "waic"))
```

```{r}
#| label: roarpa-compare-loo
#| echo: true

loo_compare(roarpa_lcdm, roarpa_dina, criterion = "loo")

loo_compare(roarpa_lcdm, roarpa_dina, criterion = "waic")
```

:::

# Reliability

## Reliability methods

* Reporting reliability depends on how results are estimated and reported

* Reliability for:
  * Profile-level classification
  * Attribute-level classification
  * Attribute-level probability of proficiency

## Reliability with measr

```{r}
#| label: ecpe-reli
#| echo: true

reliability(ecpe_lcdm)
```

## Profile-level classification

:::{.panel-tabset}

### Profile-level probabilities

:::{.new-height style="--code-height: 460px;"}
```{r}
#| label: ecpe-class-prob
#| echo: true

measr_extract(ecpe_lcdm, "class_prob")
```

:::

### Profile-level classifications

:::{.new-height style="--code-height: 460px;"}
```{r}
#| label: ecpe-class
#| echo: true

measr_extract(ecpe_lcdm, "class_prob") |>
  pivot_longer(-resp_id, names_to = "profile", values_to = "prob") |>
  slice_max(order_by = prob, by = resp_id)
```

:::

### Profile reliability

```{r}
#| label: ecpe-class-reli
#| echo: true

measr_extract(ecpe_lcdm, "pattern_reliability")
```

* Estimating classification consistency and accuracy for cognitive diagnostic assessment ([Cui et al., 2012](https://doi.org/10.1111/j.1745-3984.2011.00158.x))

:::

## Attribute-level classification

:::{.panel-tabset}

### Attribute-level probabilities

:::{.new-height style="--code-height: 460px;"}
```{r}
#| label: ecpe-att-prob
#| echo: true

measr_extract(ecpe_lcdm, "attribute_prob")
```

:::

### Attribute-level classifications

:::{.new-height style="--code-height: 460px;"}
```{r}
#| label: ecpe-att-class
#| echo: true

measr_extract(ecpe_lcdm, "attribute_prob") |>
  mutate(across(where(is.double), \(x) as.integer(x > 0.5)))
```

:::

### Classification reliability

```{r}
#| label: ecpe-att-reli
#| echo: true

measr_extract(ecpe_lcdm, "classification_reliability")
```

* Measures of agreement to assess attribute-level classification accuracy and consistency for cognitive diagnostic assessments ([Johnson & Sinharay, 2018](https://doi.org/10.1111/jedm.12196))

:::

## Attribute-level probabilities

:::{.panel-tabset}

### Attribute-level probabilities

:::{.new-height style="--code-height: 460px;"}
```{r}
#| label: ecpe-att-prob
#| echo: true
```

:::

### Probability reliability

```{r}
#| label: ecpe-prob-reli
#| echo: true

measr_extract(ecpe_lcdm, "probability_reliability")
```

* The reliability of the posterior probability of skill attainment in diagnostic classification models ([Johnson & Sinharay, 2020](https://doi.org/10.3102/1076998619864550))

:::

## Exercise 5 {.exercise}

* Examine the attribute classification indices for the ROAR-PA LCDM.

* Which attributes are the most and least reliable?

{{< countdown "4:00" bottom="15%" >}}

## {.exercise data-menu-title="Solution 5"}

```{r}
#| label: roarpa-lcdm-reli
#| echo: true

measr_extract(roarpa_lcdm, "classification_reliability")
```

# {.closing data-menu-title="Closing" style="top: 0px;"}

<br>

:::{.end-title color="white" font-size="200%"}
Evaluating diagnostic classification models
:::

:::{.end-subtitle}
With Stan and measr
:::

:::{.center}
<https://learn.r-dcm.org>
:::
