---
title: Diagnostic structural <br> models
author: W. Jake Thompson, Ph.D.
html-table-processing: none
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(ggmeasr)
library(ggdist)
library(magick)
library(distributional)
library(knitr)
library(measr)
library(here)
library(fs)
library(downlit)
library(countdown)
library(cmdstanr)
library(gt)
library(dcmdata)

set_theme(plot_margin = margin(5, 0, 0, 0))

msr_colors <- c("#8ECAE6", "#023047", "#D7263D")

# gt theme ---------------------------------------------------------------------
gt_theme_rdcm <- function(gt_dat, bg_color = "#FFFFFF", ...) {
  gt_dat |>
    gt::opt_all_caps() |>
    gt::opt_table_font(
      font = list(gt::google_font("Open Sans"), gt::default_fonts())
    ) |>
    gt::tab_options(
      column_labels.background.color = bg_color,
      column_labels.border.top.width = gt::px(3),
      column_labels.border.top.color = bg_color,
      table.border.top.width = gt::px(1),
      table.border.bottom.width = gt::px(1),
      heading.align = "left",
      heading.background.color = bg_color,
      source_notes.padding = gt::px(10),
      source_notes.border.lr.width = gt::px(0),
      source_notes.font.size = 12,
      source_notes.background.color = bg_color,
      table.font.size = 16,
      data_row.padding = gt::px(3),
      table.background.color = bg_color,
      ...
    ) |>
    gt::tab_style(
      style = gt::cell_borders(
        sides = "bottom",
        color = "transparent",
        weight = gt::px(2)
      ),
      locations = gt::cells_body(
        columns = gt::everything(),
        rows = nrow(gt_dat$`_data`)
      )
    ) |>
    gt::tab_style(
      style = list(
        gt::cell_text(
          weight = "bold",
          align = "center",
          v_align = "middle",
          color = msr_colors[2]
        ),
        gt::cell_borders(
          sides = "bottom",
          color = msr_colors[2],
          weight = gt::px(3)
        )
      ),
      locations = gt::cells_column_labels(gt::everything())
    ) |>
    gt::tab_style(
      style = gt::cell_fill(color = bg_color),
      locations = list(gt::cells_body(
        columns = gt::everything(),
        rows = gt::everything()
      ))
    )
}
```

## What is a structural model?

* The measurement model controls how the attributes relate to items

* The structural model controls how the attributes relate to each other
  * Independent
  * Correlated
  * Dependencies

## Popular structural models

* Loglinear ([Rupp et al., 2010](https://www.amazon.com/Diagnostic-Measurement-Applications-Methodology-Sciences/dp/1606235273))
  * Unconstrained
  * Independent

* Dependencies or hierarchies
  * Hierarchical DCM (HDCM; [Templin & Bradshaw, 2014](https://doi.org/10.1007/s11336-013-9362-0))
  * Bayesian Network (BayesNet; [Hu & Templin](https://doi.org/10.1080/00273171.2019.1632165))

# Loglinear structural model

## Loglinear model specification

$$
\mu_c = \color{#D55E00}{\sum_{a=1}^A\gamma_{1,(a)}\alpha_{ca}} + \color{#009E73}{\sum_{a=1}^{A-1}\sum_{a^\prime=a+1}^A\gamma_{2,(a,a^\prime)}\alpha_{ca}\alpha_{ca^\prime}} + ... + \color{#56B4E9}{\gamma_{A,(a,a^\prime,...)}\prod_{a=1}^A\alpha_{ca}}
$$

:::{.fragment}
```{=html}
<span class="eqn-box2", style="background-color: #D55E00; color: white">&gamma;<sub>1,(1)</sub>: Main effect for attribute 1</span>
```
:::

:::{.fragment}
```{=html}
<span class="eqn-box2", style="background-color: #009E73; color: white">&gamma;<sub>2,(1,2)</sub>: Two-way interaction between attributes 1 and 2</span>
```
:::

:::{.fragment}
```{=html}
<span class="eqn-box2", style="background-color: #56B4E9; color: white">&gamma;<sub>A,(1,2,...)</sub>: A-way interaction between all attributes</span>
```
:::

## From kernels to class proportions

$$
\nu_c = \frac{\exp(\mu_c)}{\sum_{c=1}^C\exp(\mu_c)}
$$

* The class proportion is defined as the exponential of the class kernel divided by the sum of the exponential for all kernels

## Constraining the loglinear model

* All possible interactions: saturated model, equivalent to an unconstrained model

* Only main effects: Attributes are independent (i.e., no interactions allowed)

* Additional interactions allows for additional moments of the latent variables
  * First-order (main effects): means
  * Second-order (two-way interactions): variances
  * Third-order (three-way interactions): skewness
  * Etc.

## Defining a loglinear model

<br>

```{r}
#| label: loglinear-example
#| echo: true
#| eval: false

loglinear(max_interaction = Inf)
```

* By default, a saturated model is estimated with all interactions

## Loglinear model specification

```{r}
#| label: loglinear-ecpe
#| echo: true
#| code-line-numbers: "|5"

ecpe_loglinear <- dcm_specify(
  qmatrix = ecpe_qmatrix,
  identifier = "item_id",
  measurement_model = lcdm(),
  structural_model = loglinear(max_interaction = 2)
)
```

## Exercise 1 {.exercise}

* Open `structural-models.Rmd` and run the `setup` chunk.

* Create a model specification for the PIE field test data
  * Measurement model: LCDM
  * Structural model: Loglinear with independent attributes

{{< countdown "2:00" bottom="15%" >}}

## {.exercise data-menu-title="Solution 1"}

```{r}
#| label: pie-loglinear
#| echo: true
#| eval: false

dcm_specify(
  qmatrix = pie_ft_qmatrix,
  identifier = "task",
  measurement_model = lcdm(),
  structural_model = loglinear(max_interaction = 1)
)
```

# Attribute hierarchies

## When attributes have order

:::{.columns}

:::{.column width="45%"}
* Without any ordering, there are 2^A^ possible profiles

* Some profiles may be theoretically impossible given what we know about the domain
:::

:::{.column width="5%"}
:::

:::{.column width="50%"}
```{r}
#| label: unconstrained-profiles
#| echo: false

library(tidyverse)
library(gt)

create_profiles(
  dcm_specify(
    qmatrix = ecpe_qmatrix,
    identifier = "item_id",
    measurement_model = lcdm(),
    structural_model = unconstrained()
  )
) |>
  gt() |>
  cols_width(
    everything() ~ px(210)
  ) |>
  cols_label(
    lexical = "Lexical",
    cohesive = "Cohesive",
    morphosyntactic = "Morphosyntactic"
  ) |>
  cols_align(align = "center", columns = everything()) |>
  gt_theme_rdcm() |>
  tab_options(table.font.size = 25)
```

:::

:::


## Finding the possible profiles

:::::{.columns}

::::{.column width="45%"}
* Consider a scenario where the ECPE skills follow a defined order
  * Lexical → Cohesive → Morphosyntactic

* A respondent could be proficient on:
  * [None of the attributes]{.fragment fragment-index=1}
  * [Only lexical]{.fragment fragment-index=2}
  * [Lexical and cohesive]{.fragment fragment-index=3}
  * [All attributes]{.fragment fragment-index=4}

:::{.fragment fragment-index=5}
* All other profiles should not be possible
:::

::::

::::{.column width="5%"}
::::

::::{.column width="50%"}
:::{.r-stack}
```{r}
#| label: unconstrained-profiles-all
#| echo: false

library(tidyverse)
library(gt)

create_profiles(
  dcm_specify(
    qmatrix = ecpe_qmatrix,
    identifier = "item_id",
    measurement_model = lcdm(),
    structural_model = unconstrained()
  )
) |>
  gt() |>
  cols_width(
    everything() ~ px(210)
  ) |>
  cols_label(
    lexical = "Lexical",
    cohesive = "Cohesive",
    morphosyntactic = "Morphosyntactic"
  ) |>
  cols_align(align = "center", columns = everything()) |>
  gt_theme_rdcm() |>
  tab_options(table.font.size = 25)
```

:::{.fragment fragment-index=1}
```{r}
#| label: unconstrained-profiles-none
#| echo: false

library(tidyverse)
library(gt)

create_profiles(
  dcm_specify(
    qmatrix = ecpe_qmatrix,
    identifier = "item_id",
    measurement_model = lcdm(),
    structural_model = unconstrained()
  )
) |>
  gt() |>
  cols_width(
    everything() ~ px(210)
  ) |>
  cols_label(
    lexical = "Lexical",
    cohesive = "Cohesive",
    morphosyntactic = "Morphosyntactic"
  ) |>
  cols_align(align = "center", columns = everything()) |>
  gt_theme_rdcm() |>
  tab_options(table.font.size = 25) |>
  tab_style(
    style = list(
      cell_fill(color = palette_measr[3]),
      cell_text(color = "black", weight = "bold")
    ),
    locations = cells_body(
      columns = everything(),
      rows = c(1)
    )
  )
```

:::

:::{.fragment fragment-index=2}
```{r}
#| label: unconstrained-profiles-one
#| echo: false

library(tidyverse)
library(gt)

create_profiles(
  dcm_specify(
    qmatrix = ecpe_qmatrix,
    identifier = "item_id",
    measurement_model = lcdm(),
    structural_model = unconstrained()
  )
) |>
  gt() |>
  cols_width(
    everything() ~ px(210)
  ) |>
  cols_label(
    lexical = "Lexical",
    cohesive = "Cohesive",
    morphosyntactic = "Morphosyntactic"
  ) |>
  cols_align(align = "center", columns = everything()) |>
  gt_theme_rdcm() |>
  tab_options(table.font.size = 25) |>
  tab_style(
    style = list(
      cell_fill(color = palette_measr[3]),
      cell_text(color = "black", weight = "bold")
    ),
    locations = cells_body(
      columns = everything(),
      rows = c(1, 4)
    )
  )
```

:::

:::{.fragment fragment-index=3}
```{r}
#| label: unconstrained-profiles-two
#| echo: false

library(tidyverse)
library(gt)

create_profiles(
  dcm_specify(
    qmatrix = ecpe_qmatrix,
    identifier = "item_id",
    measurement_model = lcdm(),
    structural_model = unconstrained()
  )
) |>
  gt() |>
  cols_width(
    everything() ~ px(210)
  ) |>
  cols_label(
    lexical = "Lexical",
    cohesive = "Cohesive",
    morphosyntactic = "Morphosyntactic"
  ) |>
  cols_align(align = "center", columns = everything()) |>
  gt_theme_rdcm() |>
  tab_options(table.font.size = 25) |>
  tab_style(
    style = list(
      cell_fill(color = palette_measr[3]),
      cell_text(color = "black", weight = "bold")
    ),
    locations = cells_body(
      columns = everything(),
      rows = c(1, 4, 7)
    )
  )
```

:::

:::{.fragment fragment-index=4}
```{r}
#| label: unconstrained-profiles-three
#| echo: false

library(tidyverse)
library(gt)

create_profiles(
  dcm_specify(
    qmatrix = ecpe_qmatrix,
    identifier = "item_id",
    measurement_model = lcdm(),
    structural_model = unconstrained()
  )
) |>
  gt() |>
  cols_width(
    everything() ~ px(210)
  ) |>
  cols_label(
    lexical = "Lexical",
    cohesive = "Cohesive",
    morphosyntactic = "Morphosyntactic"
  ) |>
  cols_align(align = "center", columns = everything()) |>
  gt_theme_rdcm() |>
  tab_options(table.font.size = 25) |>
  tab_style(
    style = list(
      cell_fill(color = palette_measr[3]),
      cell_text(color = "black", weight = "bold")
    ),
    locations = cells_body(
      columns = everything(),
      rows = c(1, 4, 7, 8)
    )
  )
```

:::

:::{.fragment fragment-index=5}
```{r}
#| label: unconstrained-profiles-full
#| echo: false

library(tidyverse)
library(gt)

create_profiles(
  dcm_specify(
    qmatrix = ecpe_qmatrix,
    identifier = "item_id",
    measurement_model = lcdm(),
    structural_model = unconstrained()
  )
) |>
  gt() |>
  cols_width(
    everything() ~ px(210)
  ) |>
  cols_label(
    lexical = "Lexical",
    cohesive = "Cohesive",
    morphosyntactic = "Morphosyntactic"
  ) |>
  cols_align(align = "center", columns = everything()) |>
  gt_theme_rdcm() |>
  tab_options(table.font.size = 25) |>
  tab_style(
    style = list(
      cell_fill(color = palette_measr[3]),
      cell_text(color = "black", weight = "bold")
    ),
    locations = cells_body(
      columns = everything(),
      rows = c(1, 4, 7, 8)
    )
  ) |>
  tab_style(
    style = list(
      cell_fill(color = palette_measr[2]),
      cell_text(color = "white", weight = "bold")
    ),
    locations = cells_body(
      columns = everything(),
      rows = c(2, 3, 5, 6)
    )
  )
```

:::

:::
::::
:::::

## Structural models for hierarchies

```{r}
#| label: strc-models
#| echo: false

library(tidyverse)
library(gt)

tibble(
  model = c("HDCM", "BayesNet"),
  description = c(
    "Hard constraints on profiles based on attribute dependencies",
    "Soft constraints on profiles based on attribute dependencies"
  ),
  measr = c(
    "`hdcm()`",
    "`bayesnet()`"
  )
) |>
  gt() |>
  fmt_markdown(columns = c(measr)) |>
  cols_width(
    model ~ px(175),
    description ~ px(850),
    measr ~ px(150)
  ) |>
  cols_align(align = "center", columns = measr) |>
  gt_theme_rdcm() |>
  tab_options(table.font.size = 25)
```

## Defining a hierarchy with measr

* Attribute structure is defined using a DAG-like syntax

* Arrows indicate the direction of the relationship
  * The parent is a prerequisite for the child

:::{.fragment .fade-in-then-semi-out}
```{r}
#| label: ecpe-hierarchy-1
#| echo: true
#| eval: false

hdcm(hierarchy = "lexical -> cohesive -> morphosyntactic")
```

:::

:::{.fragment .fade-in-then-semi-out}
```{r}
#| label: ecpe-hierarchy-2
#| echo: true
#| eval: false

ecpe_structure <- "
  lexcial -> cohesive
  cohesive -> morphosyntactic
"

hdcm(hierarchy = ecpe_structure)
```

:::

:::{.fragment .fade-in-then-semi-out}
```{r}
#| label: ecpe-hierarchy-3
#| echo: true
#| eval: false

hdcm(hierarchy = "morphosyntactic <- cohesive <- lexical")
```

:::

## Exercise 2 {.exercise}

* Write the hierarchy syntax for each of these structures

<br>

::::{.columns}

:::{.column width="33%"}
```{r dev.args = list(bg = 'transparent')}
#| label: linear-structure
#| echo: false
#| message: false
#| fig-width: 3
#| fig-asp: 1

library(dagitty)
library(ggdag)
library(ggplot2)
library(dplyr)
library(patchwork)

dag_linear <- dagitty("dag { A1 -> A2 -> A3 }")
coordinates(dag_linear) <- list(
  x = c(A1 = 2, A2 = 2, A3 = 2),
  y = c(A1 = 2.6, A2 = 2, A3 = 1.4)
)

linear <- tidy_dagitty(dag_linear) |>
  as_tibble() |>
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_node(color = palette_measr[1], fill = palette_measr[1]) +
  geom_dag_text(color = "white") +
  geom_dag_edges(
    edge_color = palette_measr[1],
    edge_width = 2,
    arrow_directed = grid::arrow(length = grid::unit(5, "pt"), type = "closed")
  ) +
  expand_limits(x = c(0, 4), y = c(1.3, 2.7)) +
  theme_void() +
  theme(
    plot.background = element_rect(fill = "transparent"),
    panel.background = element_rect(fill = "transparent")
  )

linear
```

:::

:::{.column width="33%"}
```{r dev.args = list(bg = 'transparent')}
#| label: diverge-structure
#| echo: false
#| message: false
#| fig-width: 3
#| fig-asp: 1

dag_diverge <- dagitty("dag { A2 <- A1 -> A3 }")
coordinates(dag_diverge) <- list(
  x = c(A1 = 2, A2 = 1, A3 = 3),
  y = c(A1 = 2.5, A2 = 1.5, A3 = 1.5)
)

diverge <- tidy_dagitty(dag_diverge) |>
  as_tibble() |>
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_node(color = palette_measr[1], fill = palette_measr[1]) +
  geom_dag_text(color = "white") +
  geom_dag_edges(
    edge_color = palette_measr[1],
    edge_width = 2,
    arrow_directed = grid::arrow(length = grid::unit(5, "pt"), type = "closed")
  ) +
  expand_limits(x = c(0, 4), y = c(1.3, 2.7)) +
  theme_void() +
  theme(
    plot.background = element_rect(fill = "transparent"),
    panel.background = element_rect(fill = "transparent")
  )

diverge
```

:::

:::{.column width="33%"}
```{r dev.args = list(bg = 'transparent')}
#| label: complex-structure
#| echo: false
#| message: false
#| fig-width: 3
#| fig-asp: 1

dag_complex4 <- dagitty(
  "dag { A1 -> A2 -> A3 -> A5 A2 -> A4 -> A5 }"
)
coordinates(dag_complex4) <- list(
  x = c(A1 = 3, A2 = 3, A3 = 2, A4 = 4, A5 = 3),
  y = c(A1 = 2.7, A2 = 2.25, A3 = 1.75, A4 = 1.75, A5 = 1.3)
)

complex <- tidy_dagitty(dag_complex4) |>
  as_tibble() |>
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_node(color = palette_measr[1], fill = palette_measr[1]) +
  geom_dag_text(color = "white") +
  geom_dag_edges(
    edge_color = palette_measr[1],
    edge_width = 2,
    arrow_directed = grid::arrow(length = grid::unit(5, "pt"), type = "closed")
  ) +
  expand_limits(x = c(1, 5), y = c(1.1, 2.9)) +
  theme_void() +
  theme(
    plot.background = element_rect(fill = "transparent"),
    panel.background = element_rect(fill = "transparent")
  )

complex
```

:::

::::

{{< countdown "3:00" bottom="15%" >}}

## {.exercise data-menu-title="Solution 2"}

::::{.panel-tabset}

### Linear

:::{.columns}

:::{.column width="40%"}
```{r dev.args = list(bg = 'transparent')}
#| label: linear-structure
#| echo: false
#| message: false
#| fig-width: 3
#| fig-asp: 1
```

:::

:::{.column width="60%"}
```{r}
#| label: linear-solution
#| echo: true

structure <- "A1 -> A2 -> A3"
```

:::

:::

### Diverging

:::{.columns}

:::{.column width="40%"}
```{r dev.args = list(bg = 'transparent')}
#| label: diverge-structure
#| echo: false
#| message: false
#| fig-width: 3
#| fig-asp: 1
```

:::

:::{.column width="60%"}
```{r}
#| label: diverge-solution
#| echo: true

structure <- "A2 <- A1 -> A3"

structure <- "
  A1 -> A2
  A1 -> A3
"
```

:::

:::

### Complex

:::{.columns}

:::{.column width="40%"}
```{r dev.args = list(bg = 'transparent')}
#| label: complex-structure
#| echo: false
#| message: false
#| fig-width: 4
#| fig-asp: 1
```

:::

:::{.column width="60%"}
```{r}
#| label: complex-solution
#| echo: true

structure <- "
  A1 -> A2
  A2 -> A3
  A2 -> A4
  A3 -> A5
  A4 -> A5
"

structure <- "
  A1 -> A2 -> A3 -> A5
  A2 -> A4 -> A5
"

structure <- "
  A1 -> A2
  A3 <- A2 -> A4
  A3 -> A5 <- A4
"
```

:::

:::

::::

## Estimating an attribute hierarchy

:::{.panel-tabset}

### Specification

```{r}
#| label: ecpe-hdcm-spec
#| echo: true

ecpe_hdcm_spec <- dcm_specify(
  qmatrix = ecpe_qmatrix,
  identifier = "item_id",
  measurement_model = lcdm(),
  structural_model = hdcm(hierarchy = "lexical -> cohesive -> morphosyntactic")
)
```

### Estimation

```{r}
#| label: ecpe-hdcm-estimate
#| echo: true

ecpe_hdcm <- dcm_estimate(
  ecpe_hdcm_spec,
  data = ecpe_data,
  identifier = "resp_id",
  method = "pathfinder",
  backend = "cmdstanr",
  single_path_draws = 1000,
  num_paths = 10,
  draws = 2000,
  file = here("materials", "slides", "fits", "ecpe-lcdm-hdcm")
)
```

:::

## Exercise 3 {.exercise}

* We hypothesize that our PIE attributes follow a linear hierarchy

* Fit an HDCM to the PIE data where Level 1 is required before Level 2, and Level 2 is required before Level 3
  * Use MCMC as the method
  * Use `"rstan"` as the backend
  * Use 2 chains, each with 2000 total iterations, and 1000 warmup iterations

{{< countdown "5:00" bottom="15%" >}}

## {.exercise data-menu-title="Solution 3"}

:::{.panel-tabset}

### Specification

```{r}
#| label: pie-hdcm-specify
#| echo: true

pie_hdcm_spec <- dcm_specify(
  qmatrix = pie_ft_qmatrix,
  identifier = "task",
  measurement_model = lcdm(),
  structural_model = hdcm("L1 -> L2 -> L3")
)
```


### Estimation

```{r}
#| label: pie-hdcm-estimate
#| echo: true

pie_hdcm <- dcm_estimate(
  pie_hdcm_spec,
  data = pie_ft_data,
  identifier = "student",
  method = "mcmc",
  backend = "rstan",
  chains = 2,
  iter = 2000,
  warmup = 1000,
  cores = 2,
  file = here("materials", "slides", "fits", "pie-lcdm-hdcm")
)
```

:::

## Testing our hypothesized structure

* We can enforce a hierarchy on our model

* Is that structure supported by the data we gathered?

* Just like with the measurement models, we can use model comparisons to test our hypothesis
  * Fit a model with all possible profiles
  * Fit a model with only hypothesized profiles

* If structure holds, the two models should be about equal (i.e., removing profiles doesn't hurt anything)

## Evaluating attribute structures with LOO

:::{.panel-tabset}

### Saturated model

:::{.new-height style="--code-height: 460px;"}
```{r}
#| label: read-ecpe
#| echo: true

ecpe_spec <- dcm_specify(
  qmatrix = ecpe_qmatrix,
  identifier = "item_id",
  measurement_model = lcdm(),
  structural_model = unconstrained()
)

ecpe_lcdm <- dcm_estimate(
  dcm_spec = ecpe_spec,
  data = ecpe_data,
  identifier = "resp_id",
  method = "pathfinder",
  backend = "cmdstanr",
  single_path_draws = 1000,
  num_paths = 10,
  draws = 2000,
  file = here("materials", "slides", "fits", "ecpe-lcdm-uncst")
)
```

:::

### Model comparison

```{r}
#| label: ecpe-criteria

ecpe_lcdm <- add_criterion(ecpe_lcdm, criterion = c("loo", "waic"))
ecpe_hdcm <- add_criterion(ecpe_hdcm, criterion = c("loo", "waic"))
```

```{r}
#| label: ecpe-attr-comparison
#| echo: true

loo_compare(ecpe_lcdm, ecpe_hdcm)
```

:::

## Exercise 4 {.exercise}

* Compare our unconstrained model to the HDCM we just estimated

* Which model is preferred? What does that tell us about our theory?

{{< countdown "5:00" bottom="15%" >}}

## {.exercise data-menu-title="Solution 4"}

:::{.panel-tabset}

### Saturated model

```{r}
#| label: read-pie
#| echo: true

pie_lcdm_spec <- dcm_specify(
  qmatrix = pie_ft_qmatrix,
  identifier = "task",
  measurement_model = lcdm(),
  structural_model = unconstrained()
)

pie_lcdm <- dcm_estimate(
  pie_lcdm_spec,
  data = pie_ft_data,
  identifier = "student",
  method = "mcmc",
  backend = "rstan",
  chains = 2,
  iter = 2000,
  warmup = 1000,
  cores = 2,
  file = here("materials", "slides", "fits", "pie-lcdm-uncst")
)
```

### Model comparison

```{r}
#| label: pie-criteria

pie_lcdm <- add_criterion(pie_lcdm, criterion = c("loo", "waic"))
pie_hdcm <- add_criterion(pie_hdcm, criterion = c("loo", "waic"))
```

```{r}
#| label: pie-attr-comparison
#| echo: true

loo_compare(pie_lcdm, pie_hdcm)
```

:::

# {.closing data-menu-title="Closing"}

<br>

:::{.end-title color="white" font-size="200%"}
Structural models
:::

:::{.end-subtitle}
Specification and estimation
:::

:::{.center}
<https://learn.r-dcm.org>
:::
