---
title: Diagnostic measurement models
author: W. Jake Thompson, Ph.D.
---

## Diagnostic assessment items

* Can be multidimensional

* No continuum of student achievement

* Categorical constructs
  * Usually binary (e.g., master/nonmaster, proficient/not proficient)
  
## DCM measurement models {background-image="figure/backgrounds/default.png" background-size="contain"}

* Items can measure one or both attributes

* Different DCMs define &pi;<sub>ic</sub> in different ways
  * Each DCM makes different assumptions about how attributes proficiencies combine/interact to produce an item response

* Item characteristic bar charts

## Single-attribute DCM item {background-image="figure/backgrounds/default.png" background-size="contain"}

:::{.columns}
:::{.column width="30%"}
* Item measures just attribute 1

* Respondents who are proficient on attribute 1 have high probability of correct response, regardless of other attributes
:::

:::{.column width="70%"}
```{r dcm-single-att-item}
#| out-width: 100%
#| out-height: 50%
#| fig-alt: "Bar graph showing a high probability of providing a correct response when proficient on attribute 1."

icb <- tribble(
  ~profile,    ~single, ~comp, ~noncomp, ~partial, ~lcdm,
  "[0, 0, 0]",    0.15,  0.15,     0.15,     0.15,  0.15,
  "[1, 0, 0]",    0.90,  0.90,     0.15,     0.50,  0.60,
  "[0, 1, 0]",    0.15,  0.90,     0.15,     0.35,  0.40,
  "[0, 0, 1]",    0.15,  0.15,     0.15,     0.15,  0.15,
  "[1, 1, 0]",    0.90,  0.90,     0.90,     0.75,  0.90,
  "[1, 0, 1]",    0.90,  0.90,     0.15,     0.50,  0.60,
  "[0, 1, 1]",    0.15,  0.90,     0.15,     0.35,  0.40,
  "[1, 1, 1]",    0.90,  0.90,     0.90,     0.75,  0.90
) |> 
  mutate(profile = fct_inorder(profile))

icb |> 
  ggplot(aes(x = profile, y = single, fill = single > 0.5)) +
  geom_col(show.legend = FALSE) +
  expand_limits(y = c(0, 1)) +
  scale_y_percent(breaks = seq(0, 1, by = 0.2)) +
  scale_fill_manual(values = c(`TRUE` = palette_measr[4],
                               `FALSE` = palette_measr[2])) +
  labs(x = "Profile", y = "Probability of Correct Response")
```
:::
:::

## Multi-attribute items {background-image="figure/backgrounds/default.png" background-size="contain"}

* When items measure multiple attributes, what level of mastery is needed in order to provide a correct response?

* Many different types of DCMs that define this probability differently
  * Compensatory (e.g., DINO)
  * Noncompensatory (e.g., DINA)
  * Partially compensatory (e.g., C-RUM)

* General diagnostic models (e.g., LCDM)

* Each DCM makes different assumptions about how attributes proficiencies combine/interact to produce an item response

## Compensatory DCMs {background-image="figure/backgrounds/default.png" background-size="contain"}

:::{.columns}
:::{.column width="30%"}
* Item measures attributes 1 and 2

* Must be proficient in at least 1 attribute measured by the item to provide a correct response

* Deterministic inputs, noisy "or" gate (DINO; [Templin & Henson, 2006](https://doi.org/10.1037/1082-989X.11.3.287))
:::

:::{.column width="70%"}
```{r}
#| out-width: 100%
#| out-height: 50%
#| fig-alt: "Bar graph showing a high probability of providing a correct response when proficient on either attribute 1 or attribute 2."

icb |> 
  ggplot(aes(x = profile, y = comp, fill = comp > 0.5)) +
  geom_col(show.legend = FALSE) +
  expand_limits(y = c(0, 1)) +
  scale_y_percent(breaks = seq(0, 1, by = 0.2)) +
  scale_fill_manual(values = c(`TRUE` = palette_measr[4],
                               `FALSE` = palette_measr[2])) +
  labs(x = "Profile", y = "Probability of Correct Response")
```
:::
:::

## Non-compensatory DCMs {background-image="figure/backgrounds/default.png" background-size="contain"}

:::{.columns}
:::{.column width="30%"}
* Item measures attributes 1 and 2

* Must be proficient in all attributes measured by the item to provide a correct response

* Deterministic inputs, noisy "and" gate (DINA; [de la Torre & Douglas, 2004](https://doi.org/10.1007/BF02295640))
:::

:::{.column width="70%"}
```{r}
#| out-width: 100%
#| out-height: 50%
#| fig-alt: "Bar graph showing a high probability of providing a correct response when proficient on both attribute 1 and attribute 2."

icb |> 
  ggplot(aes(x = profile, y = noncomp, fill = noncomp > 0.5)) +
  geom_col(show.legend = FALSE) +
  expand_limits(y = c(0, 1)) +
  scale_y_percent(breaks = seq(0, 1, by = 0.2)) +
  scale_fill_manual(values = c(`TRUE` = palette_measr[4],
                               `FALSE` = palette_measr[2])) +
  labs(x = "Profile", y = "Probability of Correct Response")
```
:::
:::

## Partially Compensatory DCMs {background-image="figure/backgrounds/default.png" background-size="contain"}

:::{.columns}
:::{.column width="30%"}
* Separate increases for each acquired attribute

* Compensatory reparameterized unified model (C-RUM; [Hartz, 2002](https://www.proquest.com/openview/f2b96e40dc1c5aded37b703d860f7c21/1))
:::

:::{.column width="70%"}
```{r}
#| out-width: 100%
#| out-height: 50%
#| fig-alt: "Bar graph showing a high probability of providing a correct response when proficient on both attribute 1 and attribute 2 and a moderate probability when only proficient on one of the attributes."

icb |> 
  mutate(prof = case_when(partial < 0.2 ~ "FALSE",
                          partial > 0.5 ~ "TRUE",
                          TRUE ~ "Partial")) |> 
  ggplot(aes(x = profile, y = partial, fill = prof)) +
  geom_col(show.legend = FALSE) +
  expand_limits(y = c(0, 1)) +
  scale_y_percent(breaks = seq(0, 1, by = 0.2)) +
  scale_fill_manual(values = c("TRUE" = palette_measr[4],
                               "FALSE" = palette_measr[2],
                               "Partial" = "#4B3F72")) +
  labs(x = "Profile", y = "Probability of Correct Response")
```
:::
:::

## Which DCM to use? {background-image="figure/backgrounds/default.png" background-size="contain"}

* DINO, DINA, and C-RUM are just 3 of the MANY models that are available
* Each model comes with its own set of restrictions, and we typically have to specify a single model that is used for all items (software constraint)

* General form diagnostic models  
  * Flexible; can subsume other more restrictive models
  * Again, several possibilities (e.g., G-DINA, GDM, LCDM)
  
## General DCMs {background-image="figure/backgrounds/default.png" background-size="contain"}

:::{.columns}
:::{.column width="30%"}
* Different response probabilities for each class (partially compensatory)

* Log-linear cognitive diagnostic model (LCDM; [Henson et al., 2009](https://doi.org/10.1007/s11336-008-9089-5))

* This will be our focus
:::

:::{.column width="70%"}
```{r}
#| out-width: 100%
#| out-height: 50%
#| fig-alt: "Bar graph showing a high probability of providing a correct response when proficient on both attribute 1 and attribute 2 and a moderate probability when only proficient on one of the attributes."

icb |> 
  mutate(prof = case_when(lcdm < 0.2 ~ "FALSE",
                          lcdm > 0.8 ~ "TRUE",
                          TRUE ~ "Partial")) |> 
  ggplot(aes(x = profile, y = lcdm, fill = prof)) +
  geom_col(show.legend = FALSE) +
  expand_limits(y = c(0, 1)) +
  scale_y_percent(breaks = seq(0, 1, by = 0.2)) +
  scale_fill_manual(values = c("TRUE" = palette_measr[4],
                               "FALSE" = palette_measr[2],
                               "Partial" = "#4B3F72")) +
  labs(x = "Profile", y = "Probability of Correct Response")
```
:::
:::
