---
title: Diagnostic measurement <br> models
author: W. Jake Thompson, Ph.D.
html-table-processing: none
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(ggmeasr)
library(ggdist)
library(magick)
library(distributional)
library(knitr)
library(measr)
library(here)
library(fs)
library(downlit)
library(countdown)
library(cmdstanr)
library(gt)

set_theme(plot_margin = margin(5, 0, 0, 0))

msr_colors <- c("#8ECAE6", "#023047", "#D7263D")

# gt theme ---------------------------------------------------------------------
gt_theme_rdcm <- function(gt_dat, bg_color = "#FFFFFF", ...) {
  gt_dat |>
    gt::opt_all_caps() |>
    gt::opt_table_font(
      font = list(gt::google_font("Open Sans"), gt::default_fonts())
    ) |>
    gt::tab_options(
      column_labels.background.color = bg_color,
      column_labels.border.top.width = gt::px(3),
      column_labels.border.top.color = bg_color,
      table.border.top.width = gt::px(1),
      table.border.bottom.width = gt::px(1),
      heading.align = "left",
      heading.background.color = bg_color,
      source_notes.padding = gt::px(10),
      source_notes.border.lr.width = gt::px(0),
      source_notes.font.size = 12,
      source_notes.background.color = bg_color,
      table.font.size = 16,
      data_row.padding = gt::px(3),
      table.background.color = bg_color,
      ...
    ) |>
    gt::tab_style(
      style = gt::cell_borders(
        sides = "bottom",
        color = "transparent",
        weight = gt::px(2)
      ),
      locations = gt::cells_body(
        columns = gt::everything(),
        rows = nrow(gt_dat$`_data`)
      )
    ) |>
    gt::tab_style(
      style = list(
        gt::cell_text(
          weight = "bold",
          align = "center",
          v_align = "middle",
          color = msr_colors[2]
        ),
        gt::cell_borders(
          sides = "bottom",
          color = msr_colors[2],
          weight = gt::px(3)
        )
      ),
      locations = gt::cells_column_labels(gt::everything())
    ) |>
    gt::tab_style(
      style = gt::cell_fill(color = bg_color),
      locations = list(gt::cells_body(
        columns = gt::everything(),
        rows = gt::everything()
      ))
    )
}
```

# Example data

## Data for examples

* Examination for the certificate of proficiency in English (ECPE; [Templin & Hoffman, 2013](https://doi.org/10.1111/emip.12010))
  * 28 items measuring 3 total attributes
  * 2,922 respondents

* 3 attributes
  * Morphosyntactic rules
  * Cohesive rules
  * Lexical rules

## ECPE data

```{r}
#| label: ecpe-data
#| echo: true

library(dcmdata)

ecpe_data
```

## ECPE Q-matrix

```{r}
#| label: ecpe-qmatrix
#| echo: true
#| code-line-numbers: "|5|6,12|7,11"

ecpe_qmatrix
```

## Data for exercises {.exercise}

* Pathways for Instructionally Embedded Assessment (PIE; [ATLAS, 2025](https://pie.atlas4learning.org/sites/default/files/documents/resources/PIE_Assessment_Design_Development_Technical_Report.pdf))

## Exercise 1 {.exercise}

1. Download the exercise files
```{r}
#| label: download-materials
#| echo: true
#| eval: false

usethis::use_course("r-dcm/ncme2026-exercises")
```

2. Open `measurement-models.Rmd`

3. Run the `setup` chunk

4. Explore the PIE field test data and Q-matrix
    * How many items are in the data?
    * How many respondents are in the data? 
    * How many attributes are measured?
    
{{< countdown "3:00" bottom="15%" >}}

## {.exercise data-menu-title="Solution 1"}

::: {.panel-tabset}

### PIE FT data

:::{.columns}

:::{.column width="60%"}
```{r}
#| label: pie-data
#| echo: true

pie_ft_data
```

:::

:::{.column width="40%"}

* 15 items
* 172 respondents

:::

:::

### PIE Q-matrix

:::{.columns}

:::{.column width="60%"}
```{r}
#| label: pie-qmatrix
#| echo: true

pie_ft_qmatrix
```

:::

:::{.column width="40%"}

* 15 items
* 3 attributes representing levels in a learning pathway
  * Level 1 (`L1`)
  * Level 2 (`L2`)
  * Level 3 (`L3`)

:::

:::

:::

## Model specification

Specify a DCM with `dcm_specify()`<br><br>

```{r}
#| label: specify-ecpe
#| echo: true

ecpe_spec <- dcm_specify(
  qmatrix = ecpe_qmatrix, # <1>
  identifier = "item_id", # <1>
  measurement_model = lcdm(), # <2>
  structural_model = unconstrained() # <2>
)
```

1. Specify your Q-matrix, and ID column (if present)
2. Choose the measurement and structural model to estimate

## Model estimation

Estimate a DCM with `dcm_estimate()`<br><br>

```{r}
#| label: estimate-ecpe
#| echo: true

ecpe_lcdm <- dcm_estimate(
  dcm_spec = ecpe_spec, # <1>
  data = ecpe_data, # <2>
  identifier = "resp_id", # <2>
  method = "pathfinder", # <3>
  backend = "cmdstanr", # <3>
  single_path_draws = 1000, # <4>
  num_paths = 10, # <4>
  draws = 2000, # <4>
  file = here("materials", "slides", "fits", "ecpe-lcdm-uncst") # <5>
)
```

1. Define your model specification
2. Specify your data and ID column (if present)
3. Choose your estimation method and backend
4. Pass additional arguments to the backend
5. Save the model for future efficiency


## `dcm_estimate()` options

* `method`: How to estimate the model
  * Full posterior sampling with "mcmc"
  * Quicker, but more limited "optim"
  * Posterior approximation with "variational" (default) or "pathfinder"
    * Variational inference does not always converge
    * Pathfinder is only available for the cmdstanr backend

* `...`: Additional arguments that are passed to, depending on the `method` and `backend`

:::{.columns}
:::{.column width="50%"}
* 
  * `r autolink("rstan::sampling()")`
  * `r autolink("rstan::optimizing()")`
  * `r autolink("rstan::vb()")`
:::

:::{.column width="50%"}
*
  * `r autolink("cmdstanr::sample()")`
  * `r autolink("cmdstanr::optimize()")`
  * `r autolink("cmdstanr::variational()")`
  * `r autolink("cmdstanr::pathfinder()")`
:::

:::  

# DCM item models

## Diagnostic assessment items

* Can be multidimensional

* No continuum of student achievement

* Categorical constructs
  * Usually binary (e.g., master/nonmaster, proficient/not proficient)
  
## DCM measurement models

* Items can measure one or both attributes

* Different DCMs define &pi;<sub>ic</sub> in different ways
  * Each DCM makes different assumptions about how attributes proficiencies combine/interact to produce an item response

* Item characteristic bar charts

## Single-attribute DCM item

:::{.columns}
:::{.column width="30%"}
* Item measures just attribute 1

* Respondents who are proficient on attribute 1 have high probability of correct response, regardless of other attributes
:::

:::{.column width="70%"}
```{r}
#| label: dcm-single-att-item
#| out-width: 100%
#| out-height: 50%
#| fig-alt: "Bar graph showing a high probability of providing a correct response when proficient on attribute 1."

icb <- tribble(
  ~profile    , ~single , ~comp , ~noncomp , ~partial , ~lcdm ,
  "[0, 0, 0]" , 0.15    , 0.15  , 0.15     , 0.15     , 0.15  ,
  "[1, 0, 0]" , 0.90    , 0.90  , 0.15     , 0.50     , 0.60  ,
  "[0, 1, 0]" , 0.15    , 0.90  , 0.15     , 0.35     , 0.40  ,
  "[0, 0, 1]" , 0.15    , 0.15  , 0.15     , 0.15     , 0.15  ,
  "[1, 1, 0]" , 0.90    , 0.90  , 0.90     , 0.75     , 0.90  ,
  "[1, 0, 1]" , 0.90    , 0.90  , 0.15     , 0.50     , 0.60  ,
  "[0, 1, 1]" , 0.15    , 0.90  , 0.15     , 0.35     , 0.40  ,
  "[1, 1, 1]" , 0.90    , 0.90  , 0.90     , 0.75     , 0.90
) |>
  mutate(profile = fct_inorder(profile))

icb |>
  ggplot(aes(x = profile, y = single, fill = single > 0.5)) +
  geom_col(show.legend = FALSE) +
  expand_limits(y = c(0, 1)) +
  scale_y_percent(breaks = seq(0, 1, by = 0.2)) +
  scale_fill_manual(
    values = c(`TRUE` = palette_measr[4], `FALSE` = palette_measr[2])
  ) +
  labs(x = "Profile", y = "Probability of Correct Response")
```

:::
:::

## Multi-attribute items

* When items measure multiple attributes, what level of mastery is needed in order to provide a correct response?

* Many different types of DCMs that define this probability differently
  * Compensatory (e.g., DINO)
  * Noncompensatory (e.g., DINA)
  * Partially compensatory (e.g., C-RUM)

* General diagnostic models (e.g., LCDM)

* Each DCM makes different assumptions about how attributes proficiencies combine/interact to produce an item response

## Compensatory DCMs

:::{.columns}
:::{.column width="30%"}
* Item measures attributes 1 and 2

* Must be proficient in at least 1 attribute measured by the item to provide a correct response

* Deterministic inputs, noisy "or" gate (DINO; [Templin & Henson, 2006](https://doi.org/10.1037/1082-989X.11.3.287))
:::

:::{.column width="70%"}
```{r}
#| label: dcm-compensatory-item
#| out-width: 100%
#| out-height: 50%
#| fig-alt: "Bar graph showing a high probability of providing a correct response when proficient on either attribute 1 or attribute 2."

icb |>
  ggplot(aes(x = profile, y = comp, fill = comp > 0.5)) +
  geom_col(show.legend = FALSE) +
  expand_limits(y = c(0, 1)) +
  scale_y_percent(breaks = seq(0, 1, by = 0.2)) +
  scale_fill_manual(
    values = c(
      `TRUE` = palette_measr[4],
      `FALSE` = palette_measr[2]
    )
  ) +
  labs(x = "Profile", y = "Probability of Correct Response")
```

:::
:::

## Non-compensatory DCMs

:::{.columns}
:::{.column width="30%"}
* Item measures attributes 1 and 2

* Must be proficient in all attributes measured by the item to provide a correct response

* Deterministic inputs, noisy "and" gate (DINA; [de la Torre & Douglas, 2004](https://doi.org/10.1007/BF02295640))
:::

:::{.column width="70%"}
```{r}
#| label: dcm-noncompensatory-item
#| out-width: 100%
#| out-height: 50%
#| fig-alt: "Bar graph showing a high probability of providing a correct response when proficient on both attribute 1 and attribute 2."

icb |>
  ggplot(aes(x = profile, y = noncomp, fill = noncomp > 0.5)) +
  geom_col(show.legend = FALSE) +
  expand_limits(y = c(0, 1)) +
  scale_y_percent(breaks = seq(0, 1, by = 0.2)) +
  scale_fill_manual(
    values = c(`TRUE` = palette_measr[4], `FALSE` = palette_measr[2])
  ) +
  labs(x = "Profile", y = "Probability of Correct Response")
```

:::
:::

## Partially Compensatory DCMs

:::{.columns}
:::{.column width="30%"}
* Separate increases for each acquired attribute

* Compensatory reparameterized unified model (C-RUM; [Hartz, 2002](https://www.proquest.com/openview/f2b96e40dc1c5aded37b703d860f7c21/1))
:::

:::{.column width="70%"}
```{r}
#| label: dcm-partial-compensatory-item
#| out-width: 100%
#| out-height: 50%
#| fig-alt: "Bar graph showing a high probability of providing a correct response when proficient on both attribute 1 and attribute 2 and a moderate probability when only proficient on one of the attributes."

icb |>
  mutate(
    prof = case_when(
      partial < 0.2 ~ "FALSE",
      partial > 0.5 ~ "TRUE",
      TRUE ~ "Partial"
    )
  ) |>
  ggplot(aes(x = profile, y = partial, fill = prof)) +
  geom_col(show.legend = FALSE) +
  expand_limits(y = c(0, 1)) +
  scale_y_percent(breaks = seq(0, 1, by = 0.2)) +
  scale_fill_manual(
    values = c(
      "TRUE" = palette_measr[4],
      "FALSE" = palette_measr[2],
      "Partial" = "#4B3F72"
    )
  ) +
  labs(x = "Profile", y = "Probability of Correct Response")
```

:::
:::

## Which DCM to use?

* DINO, DINA, and C-RUM are just three of the MANY models that are available
* Each model comes with its own set of restrictions, and we typically have to specify a single model that is used for all items (software constraint)

* General form diagnostic models  
  * Flexible; can subsume other more restrictive models
  * Again, several possibilities (e.g., G-DINA, GDM, LCDM)
  
## General DCMs

:::{.columns}
:::{.column width="30%"}
* Different response probabilities for each class (partially compensatory)

* Log-linear cognitive diagnostic model (LCDM; [Henson et al., 2009](https://doi.org/10.1007/s11336-008-9089-5))

:::

:::{.column width="70%"}
```{r}
#| label: dcm-general-item
#| out-width: 100%
#| out-height: 50%
#| fig-alt: "Bar graph showing a high probability of providing a correct response when proficient on both attribute 1 and attribute 2 and a moderate probability when only proficient on one of the attributes."

icb |>
  mutate(
    prof = case_when(
      lcdm < 0.2 ~ "FALSE",
      lcdm > 0.8 ~ "TRUE",
      TRUE ~ "Partial"
    )
  ) |>
  ggplot(aes(x = profile, y = lcdm, fill = prof)) +
  geom_col(show.legend = FALSE) +
  expand_limits(y = c(0, 1)) +
  scale_y_percent(breaks = seq(0, 1, by = 0.2)) +
  scale_fill_manual(
    values = c(
      "TRUE" = palette_measr[4],
      "FALSE" = palette_measr[2],
      "Partial" = "#4B3F72"
    )
  ) +
  labs(x = "Profile", y = "Probability of Correct Response")
```

:::
:::

## The LCDM as a general DCM

* So called "general" DCM because the LCDM subsumes other DCMs

* Constraints on item parameters make LCDM equivalent to other DCMs (e.g., DINA and DINO)
  * DINA
    * Only the intercept and highest-order interaction are non-0
  * DINO
    * All main effects are equal
    * All two-way interactions are -1 $\times$ main effect
    * All three-way interactions are -1 $\times$ two-way interaction (i.e., equal to main effects)
    * Etc.
  * C-RUM
    * Only the intercept and main effects are non-0 (i.e., interactions are not estimated)
  * Interactive Shiny app: <https://atlas-aai.shinyapps.io/dcm-probs/>

## Measurement models with measr

We define a DCM using `dcm_specify()`<br><br>

```{r}
#| label: specify-lcdm-ecpe
#| echo: true
#| code-line-numbers: "|4"

ecpe_spec <- dcm_specify(
  qmatrix = ecpe_qmatrix,
  identifier = "item_id",
  measurement_model = lcdm(),
  structural_model = unconstrained()
)
```

## Supported models

<br>

```{r}
#| label: meas-models
#| echo: false

tibble(
  model = c("LCDM", "DINA", "NIDA", "NC-RUM", "DINO", "NIDO", "C-RUM"),
  description = c(
    "General and flexible, subsumes other models",
    "All attributes must be present",
    "Attributes have multiplicative penalties equal across items",
    "Attributes have multiplicative penalites that vary across items",
    "Any one attribute must be present",
    "Attributes are additive and equal across items",
    "Attributes are additive and vary across items"
  ),
  measr = c(
    "`lcdm()`",
    "`dina()`",
    "`nida()`",
    "`ncrum()`",
    "`dino()`",
    "`nido()`",
    "`crum()`"
  )
) |>
  gt() |>
  fmt_markdown(columns = c(measr)) |>
  cols_width(
    model ~ px(175),
    description ~ px(850),
    measr ~ px(150)
  ) |>
  cols_align(align = "center", columns = measr) |>
  tab_row_group(
    label = "Non-compensatory",
    rows = 2:4
  ) |>
  tab_row_group(
    label = "Compensatory",
    rows = 5:7
  ) |>
  row_group_order(groups = c(NA, "Non-compensatory", "Compensatory")) |>
  gt_theme_measr() |>
  tab_options(table.font.size = 25) |>
  tab_style(
    style = list(
      cell_fill(color = msr_colors[2]),
      cell_text(color = "white"),
      cell_borders(weight = px(0))
    ),
    locations = cells_row_groups()
  )
```

## Choose a new measurement model

```{r}
#| label: specify-ecpe-dina
#| echo: true
#| code-line-numbers: "|4"

ecpe_dina_spec <- dcm_specify(
  qmatrix = ecpe_qmatrix,
  identifier = "item_id",
  measurement_model = dina(),
  structural_model = unconstrained()
)
```

## Exercise 2 {.exercise}

* Estimate an LCDM and a NIDA model on the PIE data

* For the LCDM:
  * Use MCMC as the method
  * Use `"rstan"` as the backend
  * Use 2 chains, each with 2000 total iterations, and 1000 warmup iterations

* For the NIDA:
  * Use variational inference as the method
  * Use `"rstan"` as the backend
  * Use the seed `19891213`
    
{{< countdown "5:00" bottom="15%" >}}

## {.exercise data-menu-title="Solution 2"}

:::{.panel-tabset}

### LCDM

```{r}
#| label: pie-lcd
#| echo: true

pie_lcdm_spec <- dcm_specify(
  qmatrix = pie_ft_qmatrix,
  identifier = "task",
  measurement_model = lcdm(),
  structural_model = unconstrained()
)

pie_lcdm <- dcm_estimate(
  pie_lcdm_spec,
  data = pie_ft_data,
  identifier = "student",
  method = "mcmc",
  backend = "rstan",
  chains = 2,
  iter = 2000,
  warmup = 1000,
  cores = 2,
  file = here("materials", "slides", "fits", "pie-lcdm-uncst")
)
```

### NIDA

```{r}
#| label: pie-nida
#| echo: true

pie_nida_spec <- dcm_specify(
  qmatrix = pie_ft_qmatrix,
  identifier = "task",
  measurement_model = nida(),
  structural_model = unconstrained()
)

pie_nida <- dcm_estimate(
  pie_nida_spec,
  data = pie_ft_data,
  identifier = "student",
  method = "variational",
  backend = "rstan",
  seed = 19891213,
  file = here("materials", "slides", "fits", "pie-nida-uncst")
)
```

:::

## Modify a measurement model

```{r}
#| label: specify-ecpe-lcdm-max1
#| echo: true
#| code-line-numbers: "|4"

ecpe_lcdm_main_spec <- dcm_specify(
  qmatrix = ecpe_qmatrix,
  identifier = "item_id",
  measurement_model = lcdm(max_interaction = 1),
  structural_model = unconstrained()
)
```

## Exercise 3 {.exercise}

* Create a specification for a new LCDM for the PIE FT data

* Limit the model to only include intercepts, main effects, and two-way interactions

How is this model different from the original LCDM?

{{< countdown "4:00" bottom="15%" >}}

## {.exercise data-menu-title="Solution 3"}

:::{.panel-tabset}

### Examine possible parameters

```{r}
#| label: pie-lcdm-parameters
#| echo: true

get_parameters(pie_lcdm_spec)
```

### Create specification

```{r}
#| label: new-pie-lcdm-spec
#| echo: true
#| eval: false
#| code-line-numbers: "|4"

dcm_specify(
  qmatrix = pie_ft_qmatrix,
  identifier = "task",
  measurement_model = lcdm(max_interaction = 2),
  structural_model = unconstrained()
)
```

Because all of the PIE items are single-attribute items, there's no difference at all!

:::

# Comparing measurement models

## Estimate a model to compare

:::{.panel-tabset}

### Specification

:::{.new-height style="--code-height: 460px;"}
```{r}
#| label: ecpe-dina-spec
#| echo: true

ecpe_dina_spec <- dcm_specify(
  qmatrix = ecpe_qmatrix,
  identifier = "item_id",
  measurement_model = dina(),
  structural_model = unconstrained()
)
```

:::

### Estimation

:::{.new-height style="--code-height: 460px;"}
```{r}
#| label: ecpe-dina-estimate
#| echo: true

ecpe_dina <- dcm_estimate(
  ecpe_dina_spec,
  data = ecpe_data,
  identifier = "resp_id",
  method = "pathfinder",
  backend = "cmdstanr",
  single_path_draws = 1000,
  num_paths = 10,
  draws = 2000,
  file = here("materials", "slides", "fits", "ecpe-dina-uncst")
)
```

:::

:::

## Compare models with LOO

```{r}
#| label: add-ecpe-criteria
#| include: false

ecpe_lcdm <- add_criterion(ecpe_lcdm, criterion = "loo")
ecpe_dina <- add_criterion(ecpe_dina, criterion = "loo")
```

```{r}
#| label: compare-models
#| echo: true

loo_compare(ecpe_lcdm, ecpe_dina)
```

* LCDM is the preferred model
  * Preferred does not imply "good"

* Difference is much larger than the standard error of the difference
  * Approximate threshold: Absolute difference is greater than 2.5 &times; standard error of the difference ([Bengio & Grandvalet, 2004](https://www.jmlr.org/papers/v5/grandvalet04a.html))

## Exercise 4 {.exercise}

* Compare the LCDM and NIDA models we estimated for the PIE FT data

* Which model is preferred?

{{< countdown "4:00" bottom="15%" >}}

## {.exercise data-menu-title="Solution 4"}

```{r}
#| label: pie-criteria

pie_lcdm <- add_criterion(pie_lcdm, criterion = c("loo", "waic"))
pie_nida <- add_criterion(pie_nida, criterion = c("loo", "waic"))
```

```{r}
#| label: pie-compare
#| echo: true

loo_compare(pie_lcdm, pie_nida)
```

# {.closing data-menu-title="Closing"}

<br>

:::{.end-title color="white" font-size="200%"}
Measurement models
:::

:::{.end-subtitle}
Specification and estimation
:::

:::{.center}
<https://learn.r-dcm.org>
:::
