---
title: Identifying model <br> misspecifications
author: W. Jake Thompson, Ph.D.
html-table-processing: none
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(ggmeasr)
library(ggdist)
library(magick)
library(distributional)
library(knitr)
library(measr)
library(here)
library(fs)
library(downlit)
library(countdown)
library(cmdstanr)
library(gt)
library(dcmdata)

set_theme(plot_margin = margin(5, 0, 0, 0))

msr_colors <- c("#8ECAE6", "#023047", "#D7263D")

# gt theme ---------------------------------------------------------------------
gt_theme_rdcm <- function(gt_dat, bg_color = "#FFFFFF", ...) {
  gt_dat |>
    gt::opt_all_caps() |>
    gt::opt_table_font(
      font = list(gt::google_font("Open Sans"), gt::default_fonts())
    ) |>
    gt::tab_options(
      column_labels.background.color = bg_color,
      column_labels.border.top.width = gt::px(3),
      column_labels.border.top.color = bg_color,
      table.border.top.width = gt::px(1),
      table.border.bottom.width = gt::px(1),
      heading.align = "left",
      heading.background.color = bg_color,
      source_notes.padding = gt::px(10),
      source_notes.border.lr.width = gt::px(0),
      source_notes.font.size = 12,
      source_notes.background.color = bg_color,
      table.font.size = 16,
      data_row.padding = gt::px(3),
      table.background.color = bg_color,
      ...
    ) |>
    gt::tab_style(
      style = gt::cell_borders(
        sides = "bottom",
        color = "transparent",
        weight = gt::px(2)
      ),
      locations = gt::cells_body(
        columns = gt::everything(),
        rows = nrow(gt_dat$`_data`)
      )
    ) |>
    gt::tab_style(
      style = list(
        gt::cell_text(
          weight = "bold",
          align = "center",
          v_align = "middle",
          color = msr_colors[2]
        ),
        gt::cell_borders(
          sides = "bottom",
          color = msr_colors[2],
          weight = gt::px(3)
        )
      ),
      locations = gt::cells_column_labels(gt::everything())
    ) |>
    gt::tab_style(
      style = gt::cell_fill(color = bg_color),
      locations = list(gt::cells_body(
        columns = gt::everything(),
        rows = gt::everything()
      ))
    )
}
```

## Types of misspecifications

* Attributes
  * How many, and which ones?

* Items
  * Alignment to attributes
  * Measurement model

# Attribute misspecifications

## Attribute identification

* Have we identified the correct number of attributes?

* Ideally, this is a content-driven process from the beginning
  * Identify the skills or constructs we want to measure
  * Develop items specifically for those skills using principles of evidence centered design

* But often we're handed data, or are attempting to retrofit a DCM to a previously developed assessment

## Too many attributes {.exercise}

* What might some indicators be that we have included too many attributes in our model?

:::{.fragment}
* Some we'll examine today:
  * Structural parameters: Are there classes that aren't observed and aren't explained by a hierarchy?
  * Attribute correlations: If attributes are very highly correlated, are they really measuring distinct skills?
:::

## Structural parameters

```{r}
#| label: read-ecpe

ecpe_spec <- dcm_specify(
  qmatrix = ecpe_qmatrix,
  identifier = "item_id",
  measurement_model = lcdm(),
  structural_model = unconstrained()
)

ecpe_lcdm <- dcm_estimate(
  dcm_spec = ecpe_spec,
  data = ecpe_data,
  identifier = "resp_id",
  method = "pathfinder",
  backend = "cmdstanr",
  single_path_draws = 1000,
  num_paths = 10,
  draws = 2000,
  file = here("materials", "slides", "fits", "ecpe-lcdm-uncst")
)

ecpe_lcdm <- add_respondent_estimates(ecpe_lcdm)
```

```{r}
#| label: ecpe-strc
#| echo: true
#| code-line-numbers: "|5,8,11,12"

measr_extract(ecpe_lcdm, "strc_param")
```

* We see respondents in all of the expected classes

## Attribute correlations

```{r}
#| label: attr_cor

attribute_corr <- function(x) {
  attr_cat <- measr_extract(x, "attribute_prob") |>
    select(-x@data$respondent_identifier) |>
    mutate(across(everything(), \(x) as.integer(x > 0.5)))

  ret_mat <- diag(nrow = ncol(attr_cat))
  rownames(ret_mat) <- colnames(attr_cat)
  colnames(ret_mat) <- colnames(attr_cat)

  att_combo <- combn(x = colnames(attr_cat), m = 2) |>
    t()

  for (i in seq_len(nrow(att_combo))) {
    suppressMessages(
      ret_mat[att_combo[i, 1], att_combo[i, 2]] <-
        ret_mat[att_combo[i, 2], att_combo[i, 1]] <-
          psych::tetrachoric(table(
            attr_cat[[att_combo[i, 1]]],
            attr_cat[[att_combo[i, 2]]]
          ))$rho
    )
  }

  ret_mat
}
```

```{r}
#| label: ecpe-attribute-correlation
#| echo: true

attribute_corr(ecpe_lcdm)
```

* We expect to see somewhat high correlations (but not perfect) due to the attribute dependencies

## Exercise 1 {.exercise}

* Open `misspecifications.Rmd` and run the `setup` chunk.

* Examine the base rates and attribute correlations for the PIE HDCM model.

* What do the findings indicate?

{{< countdown "4:00" bottom="15%" >}}

## {.exercise data-menu-title="Solution 1"}

```{r}
#| label: read-pie

pie_hdcm_spec <- dcm_specify(
  qmatrix = pie_ft_qmatrix,
  identifier = "task",
  measurement_model = lcdm(),
  structural_model = hdcm("L1 -> L2 -> L3")
)

pie_hdcm <- dcm_estimate(
  pie_hdcm_spec,
  data = pie_ft_data,
  identifier = "student",
  method = "mcmc",
  backend = "rstan",
  chains = 2,
  iter = 2000,
  warmup = 1000,
  cores = 2,
  file = here("materials", "slides", "fits", "pie-lcdm-hdcm")
)

pie_hdcm <- add_respondent_estimates(pie_hdcm)
```

:::{.panel-tabset}

### Structural parameters

```{r}
#| label: pie-strc
#| echo: true
#| code-line-numbers: "|6"

measr_extract(pie_hdcm, "strc_param")
```

* Very few students are proficient on only Level 1 skills

### Correlations

```{r}
#| label: pie-corr
#| echo: true

attribute_corr(pie_hdcm)
```

* Correlation between Levels 1 and 2 is much higher than between Levels 2 and 3

:::

## Too few attributes {.exercise}

* What might some indicators be that we have included too few attributes in our model?

:::{.fragment}
* Some we'll examine today:
  * Local independence: Item responses should be independent conditional on the respondent's profile of proficiency
:::

## Residual correlations with Yen's Q~3~

* Residual correlations should all be relatively small (e.g., <.2)

:::{.new-height style="--code-height: 420px;"}
```{r}
#| label: ecpe-yensq3
#| echo: true

yens_q3(ecpe_lcdm)
```

:::

## Visualizing Yen's Q~3~

```{r}
#| label: vis-yensq3
#| fig-asp: .4
#| fig-alt: >
#|   Histogram showing the distribution of residual correlations. All residual
#|   correlations are within the expected bounds of -.2 to .2.

ecpe_q3 <- yens_q3(ecpe_lcdm)

ggplot(ecpe_q3, aes(x = resid_corr)) +
  geom_histogram(binwidth = 0.01, boundary = 0, fill = palette_measr[1]) +
  geom_vline(
    xintercept = c(-.2, .2),
    linetype = "dashed",
    color = palette_measr[2]
  ) +
  expand_limits(x = c(-.25, .25)) +
  labs(x = "Residual correlation", y = "Item pairs")
```

## Exercise 2 {.exercise}

* Calculate Yen's Q~3~ (residual correlations) for the PIE HDCM

* What do the findings indicate?

{{< countdown "2:00" bottom="15%" >}}

## {.exercise data-menu-title="Solution 2"}

:::{.panel-tabset}

### Q~3~ distribution

```{r}
#| label: vis-yensq3-pie
#| warning: false
#| fig-asp: .4
#| fig-alt: >
#|   Histogram showing the distribution of residual correlations. Several
#    residual correlations are outside the expected bounds of -.2 to .2.

pie_q3 <- yens_q3(pie_hdcm)

ggplot(pie_q3, aes(x = resid_corr)) +
  geom_histogram(binwidth = 0.01, boundary = 0, fill = palette_measr[1]) +
  geom_vline(
    xintercept = c(-.2, .2),
    linetype = "dashed",
    color = palette_measr[2]
  ) +
  expand_limits(x = c(-.25, .25)) +
  labs(x = "Residual correlation", y = "Item pairs")
```

### Flagged pairs

```{r}
#| label: pie-q3
#| echo: true

yens_q3(pie_hdcm) |>
  filter(flag)
```

:::

# Item misspecifications

## Diagnosing item-level problems

* [Item-level fit](/materials/slides/01-introdcm-04-evaluation-slides.html#item-level-fit) with posterior predictive model checks
  * *p*-values
  * Conditional probabilities
  * Odds ratios

* Item discrimination

* Q-matrix validation

## Item discrimination

* Evaluates how well an item is able to distinguish between attribute profiles
  * Item- and test-level discrimination ([Henson & Douglas, 2005](https://doi.org/10.1177/0146621604272623))
  * Attribute-level discrimination ([Henson et al., 2008](https://doi.org/10.1177/0146621607302478))

* Higher is better, but no clear thresholds as values are dependent on the number of attributes, model that was estimated, etc.

* Low values may indicate an item is not measuring what we intend (or not measuring well)

## CDI with measr

```{r}
#| label: ecpe-cdi
#| echo: true
#| code-line-numbers: "|8|22"

cdi(ecpe_lcdm)
```

## Exercise 3 {.exercise}

* Calculate the CDI for the PIE HDCM model

* What conclusions or recommendations might we make based on these results?

{{< countdown "2:00" bottom="15%" >}}

## {.exercise data-menu-title="Solution 3"}

```{r}
#| label: pie-cdi
#| echo: true

cdi(pie_hdcm)
```

## Q-matrix validation

* The Q-matrix defines which items measure each attribute
  * Is the Q-matrix correct?
  * Are there other Q-matrices that might better explain the observed data

* measr implements a general Q-matrix validation measure from [de la Torre & Chiu (2016)](https://doi.org/10.1007/s11336-015-9467-8)

* Assumes that the set of attributes is correct
  * Does not evaluate whether additional, fewer, or different attributes would be appropriate

## Q-matrix validation with measr

```{r}
#| label: ecpe-qmatrix-validation
#| echo: true
#| code-line-numbers: "|13"

qmatrix_validation(ecpe_lcdm)
```

## Exercise 4 {.exercise}

* Check the Q-matrix for the PIE HDCM

* Are there any items that may measure additional pathway levels?

{{< countdown "2:00" bottom="15%" >}}

## {.exercise data-menu-title="Solution 4"}

```{r}
#| label: pie-qmatrix-validation
#| echo: true

qmatrix_validation(pie_hdcm)
```

# {.closing data-menu-title="Closing"}

<br>

:::{.end-title color="white" font-size="200%"}
Model misspecifications
:::

:::{.end-subtitle}
Attributes and items
:::

:::{.center}
<https://learn.r-dcm.org>
:::
